---
title: "T1D - TEDDY Project"
author: "Ryan Summers"
date: "2026-01-06"
output:
  word_document: default
  pdf_document: default
  html_document:
    css: style.css
bibliography: t1d.bib
csl: "apa-5th-edition.csl"
nocite: '@*'
---

```{r setup, include=FALSE}
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(pwr)
library(ssize)
library(samr)
library(parallel)
library(BiocParallel)
library(impute)
library(limma)
library(qvalue)
library(gtools)
library(multtest)
library(GEOquery)
library(Biobase)
library(data.table)
library(SummarizedExperiment)
library(variancePartition)
library(minfi)
library(R.utils)
library(stringr)
library(methylKit)
library(shinyMethyl)
library(minfi)
library(bumphunter)
library(qs)
library(qs2)
library(lme4)
library(sesame)
library(sva)
library(EmpiricalBrownsMethod)
library(clusterProfiler)
library(patchwork)
library(ggmanh)
library(survival)
library(lme4)
library(SummarizedExperiment)
library(variancePartition)
library(lmerTest)
library(future.apply)
library(data.table)
```

Scripts for installing multiple packages
```{r}
# quick installation script
BiocManager::install(c(
  "SummarizedExperiment",
  "BiocParallel",
  "variancePartition"))

install.packages(c(
  "", ""))
```

## Data Storage and Loading

```{r}
# TEMP for local testing
# setwd(paste0(Sys.getenv("RKJCOLLAB"), "/Maternal_Protection"))
# matrix_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata"
# sample_ids_path <- "data/teddy/TEDDYarrays/samples-for-analysis.csv"
# pheno_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv"
# model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntAndSlopeModel.R"
# # model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntModel.R"
# cell_types_path <- "data/teddy/TEDDYarrays/cell proportions/cell_counts_rescaled.csv"
# cov_list <- "factor(gender),new_CD8T,new_CD4T,new_NK,new_Bcell,new_Mono"
# # cov_list <- "factor(gender)"
# cov_list <- unlist(strsplit(cov_list, ","))
```


```{r}
# # load in M-value matrix
# M_read <- load("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata")
# 
# # save file using serialization - quicker
# qsave(mSet.norm.mean.plate_row, "/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", preset = "fast", nthreads = 6)
```

Load data from personal external drive
```{r}
# # load in qs file
# M.matrix <- qread("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", nthreads = 8)
# 
# qsave(M.matrix, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/M.matrix.qs", nthreads = 6)
```

850K Annotation
```{r}
# Load DNAm annotation
load(file="/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/annotation.850K.Rdata")
```


## Phenotype & M-Matrix Data Management & Cleanup

```{r}
# read in phenotype
pheno <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv")
```


```{r}
# Load list of IDs to keep in analysis
sample_ids_df <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/samples-for-analysis.csv")
sample_ids <- sample_ids_df$rgName
```


```{r}
# Load cell type proportions
cell_types <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/cell_counts_rescaled.csv")
```


```{r}
# Filter matrix to just samples for analysis
matrix.filt <- M.matrix[, which(colnames(M.matrix) %in% sample_ids)]

# save matrix file to local
qsave(matrix.filt, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs", nthreads = 6)
```

### Load M-Matrix

```{r}
# load matrix
matrix.filt <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs", nthreads = 6)

# # save into new qs2 package
# qs_save(matrix.filt, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs2", nthreads = 6)
```

### Load Pheno

```{r}
# Add cell types to pheno
  # TO NOTE: model should include all cell types except for neutrophils, doing
  # this essentially sets neutrophils as reference
pheno_filt_ct <- left_join(pheno_filt, cell_types,
                           by = c("Sample_Name" = "samplemaskid"))

# convert to years and center to help with different scale issue when modeling
pheno_filt_ct$age_yrs_c <-
  (pheno_filt_ct$sample_agedys / 365.25) -
  mean(pheno_filt_ct$sample_agedys / 365.25)

# scale the cell covariates
covars <- c("new_Bcell", "new_CD4T", "new_CD8T",
            "new_Mono", "new_NK")

pheno_filt_ct[covars] <- scale(pheno_filt_ct[covars])

# save to csv
write.csv(pheno_filt_ct, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_scaled.csv")
qsave(pheno_filt_ct, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_filt_ct.qs", nthreads = 6)
```


```{r}
# load in original pheno file 
pheno_filt_ct_unscaled <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_filt_ct.qs", nthreads = 6)

# pheno_filt_ct_unscaled <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno.csv",header=T)
```

```{r}
# convert age to years for scalability in model running
pheno_filt_ct_unscaled$age_yrs_c <-
  (pheno_filt_ct_unscaled$sample_agedys / 365.25) -
  mean(pheno_filt_ct_unscaled$sample_agedys / 365.25)

# convert age to years for scalability in modeul running
pheno_filt_ct_unscaled$age_yrs <- pheno_filt_ct_unscaled$sample_agedys / 365.25

# # save to csv
# write.csv(pheno_filt_ct_unscaled, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno.csv")
```


```{r}
# load in pheno scaled file 
pheno_filt_ct_scaled <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_scaled.csv")
```

```{r}
# convert needed variables to factors
pheno_filt_ct_unscaled <- pheno_filt_ct_unscaled %>% 
  mutate(across(c(maskid, cc, gender, CaseControl), as.factor))


pheno_filt_ct_scaled <- pheno_filt_ct_scaled %>% 
  mutate(across(c(maskid, cc, gender, CaseControl), as.factor))
```


## Exploratory Data Analysis (EDA)

```{r}
# number of patients
length(unique(pheno_filt_ct_unscaled$maskid))
```

```{r}
# missing value percentages
pheno_filt_ct_scaled %>%
  ungroup() %>% 
  miss_var_summary()

```


```{r}
# measurement stats
pheno_filt_ct %>% 
  group_by(maskid) %>%
  summarise(n_measurements = n()) -> mnts_per_id 

mnts_per_id %>%   
  summarise(avg = mean(n_measurements),
            min = min(n_measurements),
            med = median(n_measurements),
            max = max(n_measurements))

hist(mnts_per_id$n_measurements, breaks = 12, freq = FALSE)
```

```{r}
# determine number of subjects that would give "stable" estimates
subj <- pheno_filt_ct_scaled %>%
  group_by(maskid) %>%
  summarise(
    n_obs = n(),
    n_age_unique = n_distinct(sample_agedys), 
    .groups = "drop")

round(c(
  n_subjects = nrow(subj),
  n_subjects_slope_ok = sum(subj$n_age_unique >= 2),
  n_subjects_slope_better = sum(subj$n_age_unique >= 3),
  pct_slope_ok = mean(subj$n_age_unique >= 2),
  pct_slope_better = mean(subj$n_age_unique >= 3)), 3)

unique(lme_results$error_message)

```

```{r}
# check that whole-blood samples have variability
covs <- c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sds <- sapply(pheno_filt_ct_unscaled[, covs], sd, na.rm=TRUE)
which(sds == 0 | is.na(sds))
colSums(!is.finite(as.matrix(pheno_filt_ct_unscaled[, covs])))

```

```{r}
vars <- c("age_yrs_c","new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sapply(pheno_filt_ct_scaled[, vars], function(x) sum(!is.finite(x)))

sapply(pheno_filt_ct_unscaled[, c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")],
       function(x) sd(x, na.rm = TRUE))

```

```{r}
# Confirm sample ID matching across the entire dataset
sid <- as.character(pheno_filt_ct_scaled$rgName)
mean(sid %in% colnames(matrix.filt))       # should be 1
sum(!(sid %in% colnames(matrix.filt)))     # should be 0

```

Sanity check: confirm 1:1 sets
* is there a match for each case indicator? Other words, a case & control for each indicator

```{r}
check_sets <- pheno_filt_ct_scaled %>%
  dplyr::distinct(case_ind, maskid, outcome, case_endptage) %>%
  group_by(case_ind) %>%
  summarise(
    n_ids = n_distinct(maskid),
    n_cases = sum(outcome == 1),
    n_ctrls = sum(outcome == 0),
    endpt_unique = n_distinct(case_endptage),
    .groups = "drop")

table(check_sets$n_ids)
table(check_sets$n_cases)
table(check_sets$endpt_unique)

bad_sets <- check_sets %>% 
  filter(n_ids != 2 | n_cases != 1 | endpt_unique != 1)
bad_sets

```


### Spaghetti plots

```{r}
getSpagPlot <- function(CpG, mvals, pheno, anno) {
  # error handling
  stopifnot(CpG %in% rownames(mvals))
  stopifnot(CpG %in% rownames(anno))
  stopifnot("rgName" %in% names(pheno))
  stopifnot(all(c("sample_agedys","maskid") %in% names(pheno)))

  # match matrix column names by rgName in pheno
  samp <- intersect(colnames(mvals), pheno$rgName)
  if (length(samp) == 0) 
    stop("No overlap between colnames(matrix) and pheno$rgName")

  # build plotting data
  toPlot <- pheno[match(samp, pheno$rgName), , drop = F] # match sample names to pheno
  toPlot$CpG_value <- as.numeric(mvals[CpG, samp]) # match CpG & sample name to matrix

  # annotation for title
  a <- anno[CpG, , drop = F]
  title <- paste0(
    CpG, " (", a$UCSC_RefGene_Name, ") ",
    a$chr, ": ", prettyNum(a$pos, big.mark = ","))

  ggplot(toPlot, aes(x = sample_agedys, y = CpG_value, group = maskid)) +
    geom_line(alpha = 0.4) +
    theme_bw() +
    ylab(paste0(CpG, " (Beta value)")) +
    labs(title = title) +
    theme(plot.title = element_text(size = 7))
}



#plots <- lapply(df$CpG, getSpagPlot, toPlot = toPlot, anno = anno)
p <- getSpagPlot("cg10631356", 
                 mvals = matrix.filt, 
                 pheno = pheno_filt_ct_unscaled, 
                 anno = anno)
p

```


Function to identify specific CpGs and provide pheno information from the CpG
```{r}
cpg_report <- function(
    CpG, # CpG(s) of interest
    mvals, # M-value matrix
    pheno, # phenotype data set
    anno, # annotation dataset - 850k
    pheno_id = "rgName", # column in pheno file that matches matrix columns
    outlier_method = c("iqr","z"), # outlier calculation
    z_thresh = 3, # number of deviations away from mean for z-test 
    DNAm_values = c("M", "beta")  # return M- or beta-values
    ) 
  {
  # enforce
  outlier_method <- match.arg(outlier_method)
  DNAm_values  <- match.arg(DNAm_values)
  
  # M to Beta function
  m_to_beta <- function(M) {
    p <- 2^M
    p / (p + 1)
  }
  
  # error handling 
  if (!CpG %in% rownames(anno)) stop("CpG not in rownames(anno): ", CpG)
  
  cpg_in_rows <- CpG %in% rownames(mvals)
  cpg_in_cols <- CpG %in% colnames(mvals)
  if (!cpg_in_rows && !cpg_in_cols) stop("CpG not found in mvals: ", CpG)
  
  if (!pheno_id %in% names(pheno)) stop("pheno_id column not in pheno: ", pheno_id)
  
  # extract values from M-Matrix 
  vals <- if (cpg_in_rows) mvals[CpG, ] else mvals[, CpG]
  vals <- as.numeric(vals)
  names(vals) <- if (cpg_in_rows) colnames(mvals) else rownames(mvals)
  
  # convert M -> Beta if chosen
  if (DNAm_values == "beta") {
    vals <- m_to_beta(vals)
  }
  
  # match to pheno dataset 
  keep <- intersect(names(vals), pheno[[pheno_id]])
  df <- pheno[match(keep, pheno[[pheno_id]]), , drop = F]
  df$CpG_value <- vals[keep]
  
  # dentify outliers 
  x <- df$CpG_value
  if (outlier_method == "iqr") {
    q1 <- quantile(x, 0.25, na.rm = T)
    q3 <- quantile(x, 0.75, na.rm = T)
    iqr <- q3 - q1
    lo <- q1 - 1.5 * iqr
    hi <- q3 + 1.5 * iqr
    df$is_outlier <- x < lo | x > hi
    df$outlier_score <- NA_real_
    outlier_rule <- list(method="iqr", lo=lo, hi=hi, q1=q1, q3=q3, iqr=iqr)
  } else {
    z <- (x - mean(x, na.rm = T)) / sd(x, na.rm = T)
    df$is_outlier <- abs(z) >= z_thresh
    df$outlier_score <- z
    outlier_rule <- list(method="z", z_thresh=z_thresh)
  }
  
  # annotation  
  a <- anno[CpG, , drop = F]
  
  # summaries 
  df_sorted <- df[order(df$CpG_value), ]
  outliers <- df[df$is_outlier %in% T, , drop = F]
  
  # convert to a large list for organization
  list(
    CpG = CpG,
    DNAm_scale = DNAm_values,
    anno = a,
    outlier_rule = outlier_rule,
    summary = summary(df$CpG_value),
    n = nrow(df),
    n_outliers = nrow(outliers),
    outliers = outliers,
    lowest = head(df_sorted, 10),
    highest = head(df_sorted[rev(seq_len(nrow(df_sorted))), ], 10),
    data = df)
  }
```


```{r}
rep <- cpg_report("cg10631356", 
                  matrix.filt, 
                  pheno_filt_ct_unscaled, 
                  anno, 
                  outlier_method="z", z_thresh = 3)

# extract CpG values for each subject along w/ pheno data
rep$data

# get outliers
rep$outliers
```



## Sample Modeling Runs

```{r}
# test a small sample
set.seed(2522)
idx <- sample(nrow(matrix.filt), 500)
matrix.sample <- matrix.filt[idx, , drop = F]

pheno_ordered <- pheno_filt_ct_unscaled[match(colnames(matrix.sample), 
                                     pheno_filt_ct_unscaled$rgName), ]
stopifnot(all(!is.na(pheno_ordered$rgName)))
```

Testing singular CpGs produced from lmeR model w/bobyqa optimizer
```{r}
cpgs_to_test <- lmeR_bobyqa_results[lmeR_bobyqa_results$is_singular==TRUE,]$CpG
matrix.sample <- matrix.filt[cpgs_to_test, , drop = F]

pheno_ordered <- pheno_filt_ct_unscaled[match(colnames(matrix.sample), 
                                     pheno_filt_ct_unscaled$rgName), ]
stopifnot(all(!is.na(pheno_ordered$rgName)))
```



### lmeR Run

```{r}
# fit model
t0 <- Sys.time()

# testing different optimizer settings
ctrl <- lme4::lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5)) #1e5 is default

fit_cpg <- function(probe) {
  mod_function_lmer(
    probe = probe,
    matrix = matrix.sample,
    pheno = pheno_filt_ct_unscaled,
    sample_var = "rgName",    
    id_var = "maskid",           
    age_var = "age_yrs",   
    covs = c("gender", "cc", "CaseControl", "new_Bcell", "new_CD4T", 
             "new_CD8T", "new_Mono", "new_NK"),    
    return_blups = FALSE, # keep FALSE for speed
    control = ctrl,
    random_slope = age_var # 1 is intercept only
  )}

# set number of cores
future::plan(future::multicore, workers = 6)
options(future.globals.maxSize = 20 * 1024^3)  # 20 GiBb

test_probes <- rownames(matrix.sample) # dim = 790944x1407
test_res <- future_lapply(test_probes, fit_cpg)
test_res <- data.table::rbindlist(test_res, fill = TRUE)

qsave(
  test_res, 
  "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_int_only_singular.qs")

t1 <- Sys.time()

message("lmeR runtime: ", 
        round(as.numeric(difftime(t1, t0, units = "mins")), 2), 
        " minutes")
```

```{r}
nrow(test_res %>% filter(!is.na(warn_message)))
nrow(test_res %>% filter(!is.na(error_message)))
```

"Model failed to converge with max|grad| = 0.00399344 (tol = 0.002, component 1)":

That warning means lmer returned estimates, but the optimizer stopped with a gradient that’s a bit larger than the default tolerance. In plain terms:

* the algorithm got close to the optimum, but not close enough to declare full convergence.

So the results aren’t automatically “wrong,” but you should treat that CpG’s fit as potentially unreliable (especially SEs / variance components).


Eigenvalue: 
0.0345 - Substantial random-effects variability
5×10⁻⁹- Essentially zero variability

Interpretation:
There is strong subject-to-subject variability in one direction (intercept), and almost none in the other (slope-related direction).

* Eigenvalues > 0 → matrix is positive definite (good)
* Very small eigenvalues → one direction has almost no variance
*	≈ 0 eigenvalue → model is near-singular
	

```{r}
# form <- ~ age_yrs_c + gender + cc + new_Bcell + new_CD4T + new_CD8T + 
# new_Mono + new_NK + (1 | maskid)
fit_cpg <- function(probe) {
  mod_function(
    probe = probe,
    matrix = matrix.filt,
    pheno = pheno_filt_ct,
    sample_var = "rgName",    
    id_var = "maskid",           
    age_var = "age_yrs_c",   
    covs = c("gender", "cc", "new_Bcell", "new_CD4T", 
             "new_CD8T", "new_Mono", "new_NK"),    
    return_blups = FALSE # keep FALSE for speed
  )}

```

Run a small parallel test first	

```{r}
future::plan(future::multicore, workers = 6)
future::plan()
options(future.globals.maxSize = 20 * 1024^3)  # 20 GiBb

system.time({
  test_probes <- rownames(matrix.sample) # dim = 790944x1407
  test_res <- future_lapply(test_probes, fit_cpg)
  test_res <- data.table::rbindlist(test_res, fill = TRUE)
  })

# save to csv
write.csv(test_res, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/test_res.csv")

# load csv
test_res <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/test_res.csv")
```
    user   system  elapsed 
8063.489   73.480 8139.559 - 2.3hrs 

Estimated time to run all CpGs on my system:
20*8139.559 = 162791.2sec ≈ 2,714min ≈ 45.2hrs ≈ 2days


```{r}
# extract errors from test run
error_cpgs <- test_res[test_res$fit_status == "error", "CpG"]$CpG

tt[error_cpgs,]
```


Methylation Values to Beta function
```{r}
m2beta <- function(m) {
  2^m / (1 + 2^m)}
```
	

## DREAM Modeling Analysis

```{r}
# now fit new summarized experiment object w/ filtered CpGs
pheno_ordered <- pheno_filt_ct[match(colnames(matrix.filt), pheno_filt_ct$rgName), ]
stopifnot(all(!is.na(pheno_ordered$rgName)))

# build the SummarizedExperiment
se.filt <- SummarizedExperiment(
  assays = list(Mvalue = matrix.sample), #[1:39548,]
  colData = as.data.frame(pheno_ordered))

```


```{r}
# formula
form <- ~ age_yrs_c + gender + new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK + (1 | maskid)
```

```{r}
# set up parallelization
n.cores <- parallel::detectCores() - 1
param <- SnowParam(5, type = "SOCK", progressbar = T)  
param <- MulticoreParam(workers = 5, progressbar = T)
```

```{r}
# fit the model - uses REML by default
system.time({
  dream.model3 <- suppressWarnings(
    dream(exprObj = assay(se.filt, "Mvalue"), 
          formula = form, 
          data = colData(se.filt), 
          ddf = "Satterthwaite", 
          BPPARAM = param))
  })

qsave(dream.model3, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream.model3.qs", preset = "fast", nthreads = 6)
```
user   system  elapsed 
39.413   10.947 1176.774 - 19.6min

Est full:
≈ 392min ≈ 6.5hrs - 6.9x faster than custom function


```{r}
dream_test3 <- topTable(dream.model3, number = Inf)

# number of significant CpGs based on adjusted p-value
nrow(subset(dream_test3, adj.P.Val < 0.05))
nrow(subset(dream_test3, adj.P.Val < 0.05)) / nrow(matrix.sample)
```




Alpine ≈ 319.96min ≈ 5.3hrs
```{r}
# load Alpine dream model
dream.model <- qs_read("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream.model.qs2",
        nthreads = 6)
```

```{r}
# load in dream results
dream_results <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream_res.qs", nthreads = 6)
```


## Dream Model Diagnostics

```{r}
# load in DREAM results
#dream_results <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream_res.csv")
dream.model <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream.model.qs", 
                       nthreads = 6)
```

```{r}
# use topTable to view results
dream_results <- topTable(dream.model, number = Inf)

# save by serialization
qsave(dream_results, "dream_results.qs", nthreads=6)

# write to csv
write.csv(dream_results, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/dream_results.csv")

# number of significant CpGs after
nrow(subset(dream_res, adj.P.Val < 0.05))

tt_T1DAge <- topTable(dream.model, coef = "diseaseCase:age_years",
         number = Inf, sort.by = "B") 

sig_tt <- topTable(dream.model, coef = "diseaseCase:age_years",
         number = Inf, sort.by = "B") 
```


```{r}
all_cpgs    <- rownames(matrix.filt)
modeled_cpgs <- dream_results$X

# how many CpGs were dropped from model
length(setdiff(all_cpgs, modeled_cpgs))

# determine CpGs dropped from dream model
dropped <- setdiff(all_cpgs, modeled_cpgs)

# check variability of dropped CpGs
summary(apply(matrix.filt[dropped, ], 1, var))

```


```{r}
# number of significant CpGs based on adjusted p-value
nrow(subset(dream_results, adj.P.Val < 0.05))
nrow(subset(dream_results, adj.P.Val < 0.05)) / nrow(matrix.filt)

```

Check how many subjects only have one observation, which can be ill-posed for
random intercept/slope models
```{r}
pheno_filt_ct$maskid <- factor(pheno_filt_ct$maskid)
table_sizes <- table(pheno_filt_ct$maskid)
summary(table_sizes)
mean(table_sizes == 1)

```

Look at effect sizes

* logFC is the estimated regression coefficient = M-values
* AveExpr is the average level of M-values for CpG site across all samples
```{r}
tt <- topTable(dream.model2, coef="age_yrs_c", number=Inf)
summary(round(m2beta(abs(tt$logFC)), 4))  # for methylation, interpret appropriately
quantile(round(m2beta(abs(tt$logFC)),4), c(.5,.9,.99))

```

```{r}
# genomic inflation factor
chisq <- qchisq(1 - p, df=1)
lambda <- median(chisq, na.rm=TRUE) / qchisq(0.5, df=1)
lambda

```


The below Emprirical Bayes method has been shown to be inefficient:

"We note that an earlier version (before v1.15.5 from September 5, 2019) of dream in the
variancePartition package included an empirical Bayes step. However, after additional
testing, we removed this step so that only gene-level estimates are used in the linear mixed
model. We note that the elevated false positive rate that Yu et al. (2019) observed in simu-
lations using dream are due this this issue and has been resolved." - Hoffman et al 2020 supp
```{r}
# Empirical Bayes - pulls info across all CpGs (stabilizes variances) 
dream.model.EB <- eBayes(dream.model)

# qs methods - soon to be deprecated
qsave(dream.model.EB, "dream.model.EB.qs", preset = "fast", nthreads = 6)


# load dream object
system.time({
  dream.model.EB <- qread("dream.model.EB.qs", nthreads = 8)
  })

```


“For each CpG we attempted a random-slope model. When fits were singular or showed convergence issues, we refit using a reduced random-effects structure (uncorrelated slope or random-intercept only). CpGs were excluded only if the reduced model failed or had insufficient observations.”



```{r}
cg25271479_df <- colData(se.filt) %>% 
  as.data.frame() %>%
  mutate(methylation = assay(se.filt, "Mvalue")["cg25271479", ])

## There is no exact, universally accepted denominator df because FE estimates
## depend on RE str and var components
## Therefore, lmer() omits p-values because df are ambiguous; in practice, lmerTest (Satterthwaite) is the standard way to obtain them, especially for large-scale longitudinal omics analyses.
cg25271479.lmer.model <- lmer(
  methylation ~ sample_agedys + gender + cc + (sample_agedys | maskid),
  data = cg25271479_df)

summary(cg25271479.lmer.model)

## The REML likelihood depends on the fixed‐effects design matrix
## REML integrates out fixed effects
## If you compare models with different fixed effects, the REML likelihoods are not comparable
ranova(cg25271479.lmer.model)

```



How to see what is singular in your model
```{r}
VarCorr(cg25271479.lmer.model)
```


## lme & lmeR Alpine Model Results 

```{r}
# load in Step1B original model results
step1b_og_results <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/step1b_og_results.qs", nthreads = 6)
```

```{r}
step1b_scaled_results <- qs_read("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/Step1B_Model_scaled.qs2", nthreads = 6)
```

```{r}
# load lme modeling results from Alpine
lme_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lme_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results from Alpine
lmeR_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results using bobyqa optimizer from Alpine
lmeR_bobyqa_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_bobyqa_results.csv") %>% select(-1)
```


### Exploration Convergence Issues

```{r}
length(lme_results$var.rand.slope < 1e-8) # should matrix M-matrix
mean(lme_results$var.rand.slope < 1e-8, na.rm=TRUE)
summary(lme_results$var.rand.slope)

```

~5.3% of CpGs have effectively zero random‐slope variance
* there is no between‐subject heterogeneity in age slope

```{r}
# check data matrix (X) for redundancy - DOES NOT EVALUATE Z
mv <- ~ age_yrs_c + gender + cc + new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK
X <- model.matrix(mv, data = pheno_filt_ct_scaled)

qr(X)$rank
ncol(X)
kappa(X)

```

* Full rank -> no exact collinearity?
* How close the matrix is to being singular
  * Large singular value -> strong direction in predictor space
  * Small singular value -> nearly redundant / weakly identifiable direction
  
It does not say anything about:
* random-effects variance structure
* boundary variance estimates
* random slope identifiability
* covariance singularities

This is why kappa looks good, and we still get convergence warnings. They come from the random-effects (Z), not from X

```{r}
# check that whole-blood samples have variability
covs <- c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sds <- sapply(pheno_filt_ct_unscaled[, covs], sd, na.rm=TRUE)
which(sds == 0 | is.na(sds))
colSums(!is.finite(as.matrix(pheno_filt_ct_unscaled[, covs])))

```

```{r}
vars <- c("sample_agedys","new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sapply(pheno_filt_ct_unscaled[, vars], function(x) sum(!is.finite(x)))

sapply(pheno_filt_ct_unscaled[, c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")],
       function(x) sd(x, na.rm = TRUE))

```



### Identify CpGs that did not converge

```{r}
# identify CpGs not fit in Step1B model
step1b_bad_cpgs <- step1b_og_results %>% filter(!is.na(message))
```

```{r}
# Step1B with scaled variables - CpGs that did not fit
step1b_scaled_bad_cpgs <- step1b_scaled_results %>% filter(!is.na(message))
```

```{r}
# lme model CpGs that did not fit (my script)
lme_bad_cpgs <- lme_results %>% filter(!is.na(error_message))
```

```{r}
# lmeR model CpGs that did not fit (my script)
## the lmeR provides estimates for every CpG but comes w/ convergence warnings
lmeR_bad_cpgs <- lmeR_results %>% filter(!is.na(warn_message))
```

```{r}
# lmeR model, using bobyqa optimizer, CpGs that did not fit
## the lmeR provides estimates for every CpG but comes w/ convergence warnings
lmeR_bobyqa_bad_cpgs <- lmeR_bobyqa_results %>% filter(!is.na(warn_message))
```

```{r}
# how many of the "bad" CpGs overlap between the models
cat("Step1B, lme, lmeR, and lmeR w/bobyqa had", 
    length(intersect(step1b_bad_cpgs$CpG, lme_bad_cpgs$CpG, 
                     lmeR_bad_cpgs, lmeR_bobyqa_bad_cpgs)), 
    "overlapping CpGs that failed\n")

# unique CpGs to all models
unique_step1b <- setdiff( # finds all rows in x that aren't in y : setdiff(x,y)
  step1b_bad_cpgs$CpG,
  lme_bad_cpgs$CpG)

cat("Step1B had", length(unique_step1b), "unique CpGs that failed\n")

unique_lme <- setdiff(
  lme_bad_cpgs$CpG,
  step1b_bad_cpgs$CpG)

cat("LME had", length(unique_lme), "unique CpGs that failed\n")

```

lme runtime: 406.77 minutes ~ 6.8hrs
dream runtime: 219.8 minutes ~ 3.66hrs

Number of singular fits using lmeR
```{r}
cat("lmeR model w/ default optimization had",
    nrow(lmeR_results[lmeR_results$is_singular==TRUE,]),
    "singular fits\n\n")

cat("lmeR model w/ bobyqa optimization had",
    nrow(lmeR_bobyqa_results[lmeR_bobyqa_results$is_singular==TRUE,]),
    "singular fits\n")
```



Compare number of significant slopes across methods before multiple-testing adjustment
```{r}

# Step1B OG
cat("Step1B model had", 
    nrow(subset(step1b_og_results, fixed.sample_agedys.p.value < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(step1b_og_results, fixed.sample_agedys.p.value < 0.05)) / 
      nrow(step1b_og_results),3), "% of total CpG tested\n\n")


# lme - mine
cat("lme model had", 
    nrow(subset(lme_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lme_results, p_slope < 0.05)) / 
      nrow(lme_results),3), "% of total CpG tested\n\n")

# lmeR 
cat("lmeR model had", 
    nrow(subset(lmeR_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lmeR_results, p_slope < 0.05)) / 
      nrow(lmeR_results),3), "% of total CpG tested\n\n")

# lmeR w/ bobyqa optimization
cat("lmeR model w/bobyqa optimization had", 
    nrow(subset(lmeR_bobyqa_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lmeR_bobyqa_results, p_slope < 0.05)) / 
      nrow(lmeR_bobyqa_results),3), "% of total CpG tested\n")

```
Identify CpGs that had a significant slope but was singular - var(slope)≈0
```{r}
lmeR_bobyqa_results <- lmeR_bobyqa_results %>% 
  mutate(sig_singular = ifelse(
    is_singular == TRUE & p_slope<0.05, "Trouble", "Good"
  ))

lmeR_results <- lmeR_results %>% 
  mutate(sig_singular = ifelse(
    is_singular == TRUE & p_slope<0.05, "Trouble", "Good"
  ))


# number of CpGs with a significant slope but was singular
cat("lmeR w/default optimization had",
    nrow(lmeR_results[lmeR_results$sig_singular=="Trouble",]),
    "CpGs with a significant slope but was singular\n", 
    "(subjects are inferred to have same slope)\n\n")

cat("lmeR w/bobyqa optimization had",
    nrow(lmeR_bobyqa_results[lmeR_bobyqa_results$sig_singular=="Trouble",]),
    "CpGs with a significant slope but was singular\n", 
    "(subjects are inferred to have same slope)")
```

lmeR returned estimates for every CpG - lets compare CpG's that fit w/ lmeR with Step1B
* maybe just look at significant ones?


## Multiple Testing Adjustment


lmeR w/bobyqa optimization
```{r}
# keep CpGs w/ no singular fits
keep <- lmeR_bobyqa_results$is_singular %in% FALSE & !is.na(res$p_slope)

# Benjamini-Hochberg FDR (most common for EWAS)
res$q_slope <- NA_real_
res$q_slope[ok] <- p.adjust(res$p_slope[ok], method = "BH")
```








