---
title: "T1D - TEDDY Project"
author: "Ryan Summers"
date: "2026-01-06"
output:
  word_document: default
  pdf_document: default
  html_document:
    css: style.css
bibliography: t1d.bib
csl: "apa-5th-edition.csl"
nocite: '@*'
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# libraries
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(parallel)
library(multtest)
library(R.utils)
library(stringr)
library(qs)
library(qs2)
library(lme4)
library(patchwork)
library(lme4)
library(lmerTest)
library(future.apply)
library(data.table)
library(splines)
library(naniar)
library(clusterCrit)
library(mclust)
library(emmeans)
library(bigmemory)
library(cluster)
library(dpGMM)
library(fda)
library(segmented)

# Scripts
source("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/scripts/CpG_Report.R")
source("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/scripts/spaghetti_plot_big.R")
source("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/scripts/GMM_hist2.R")

# Functions
m2beta <- function(m) {
  2^m / (1 + 2^m)}

scale_rows <- function(x) {
  out <- t(scale(t(x), center = TRUE, scale = TRUE))
  out[is.na(out)] <- 0   # flat trajectories get sd = 0
  out
  }
```

Scripts for installing multiple packages
```{r}
# quick installation script
BiocManager::install(c(
  "SummarizedExperiment",
  "BiocParallel",
  "variancePartition"))

install.packages(c(
  "", ""))
```

## Data Storage and Loading

```{r}
# TEMP for local testing
# setwd(paste0(Sys.getenv("RKJCOLLAB"), "/Maternal_Protection"))
# matrix_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata"
# sample_ids_path <- "data/teddy/TEDDYarrays/samples-for-analysis.csv"
# pheno_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv"
# model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntAndSlopeModel.R"
# # model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntModel.R"
# cell_types_path <- "data/teddy/TEDDYarrays/cell proportions/cell_counts_rescaled.csv"
# cov_list <- "factor(gender),new_CD8T,new_CD4T,new_NK,new_Bcell,new_Mono"
# # cov_list <- "factor(gender)"
# cov_list <- unlist(strsplit(cov_list, ","))
```

Infinium MethylationEPIC v2.0 Kit
```{r}
# # load in M-value matrix 
# M_read <- load("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata")
# 
# # save file using serialization - quicker
# qsave(mSet.norm.mean.plate_row, "/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", preset = "fast", nthreads = 6)
```

Load data from personal external drive
```{r}
# # load in qs file
# M.matrix <- qread("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", nthreads = 8)
# 
# qsave(M.matrix, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/M.matrix.qs", nthreads = 6)
```


## Phenotype & M-Matrix Data Management & Cleanup

```{r}
# read in phenotype
#pheno <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv")
```


```{r}
# Load list of IDs to keep in analysis
sample_ids_df <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/samples-for-analysis.csv")
sample_ids <- sample_ids_df$rgName
```


```{r}
# Load cell type proportions
cell_types <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/cell_counts_rescaled.csv")
```


```{r}
# merge pheno type datasets
pheno_complete <- dplyr::left_join(
  pheno_filt_ct_unscaled,
  pheno2,
  by = "maskid")

# clean up duplicate columns
pheno_complete <- pheno_complete %>%
  select(-ends_with(".y")) %>%
  rename_with(~ gsub("\\.x$", "", .x), ends_with(".x"))

# # save to csv
# write.csv(pheno_complete, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_complete.csv")
```

### Load Pheno 

```{r}
pheno_complete <- suppressMessages(read_csv("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/data/processed/pheno_complete.csv") %>% select(-c(1:2)))
```


### Load 850K Annotation

```{r}
# Load DNAm annotation
anno <- readRDS("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/data/processed/anno.rds")
```

```{r, eval=FALSE}
setDT(anno, keep.rownames = "CpG")

anno <- anno[, .(
  CpG = CpG,
  UCSC_RefGene_Name,
  chr,
  pos)]

anno <- as.data.frame(anno)
rownames(anno) <- anno$CpG
anno$CpG <- NULL

saveRDS(anno, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/anno.rds", 
        compress = "xz")
```

### Load M-Matrix

```{r}
# # load matrix
# matrix.filt <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs",nthreads = 12)
# 
# matrix.full <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/M.matrix.qs", nthreads = 12)
```

```{r}
# filter out chrX CpGs 
cpg_xy <- intersect(
  rownames(anno)[anno$chr %in% c("chrX", "chrY")], rownames(matrix.filt))
matrix.filt_xy <- matrix.filt[setdiff(rownames(matrix.filt), cpg_xy), ]

matrix.filt_xy <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt_xy.qs",
      nthreads = 12)

# filter out cases from matrix
cases <- pheno_complete[pheno_complete$CaseControl == "Case", "rgName"]
matrix.ctrls <- matrix.filt_xy[, !colnames(matrix.filt_xy) %in% cases$rgName]
```

```{r}
# qsave(matrix.ctrls, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.ctrls.qs", nthreads = 12)

matrix.ctrls <- qread(
  "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.ctrls.qs",
  nthreads = 12)

#icloud
matrix.ctrls <- qread("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/data/processed/matrix.ctrls.qs", nthreads = 6)
```


Creating the m-matrix as a file-backed (hard drive) object - significantly reduces ram
```{r}
# # use an absolute path
# bp <- normalizePath(
#   "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/bigdata",
#   mustWork = FALSE)
# 
# # create directory (if not done yet)
# dir.create(bp, recursive = TRUE, showWarnings = FALSE)
# 
# # remove old files
# unlink(file.path(bp, "X.bin"))
# unlink(file.path(bp, "X.desc"))
# 
# # strip dimnames completely
# #dimnames(matrix.clean) <- NULL
# 
# # create the backing
# X_bm <- filebacked.big.matrix(
#   nrow = nrow(matrix.ctrls),
#   ncol = ncol(matrix.ctrls),
#   type = "double",
#   backingfile = "X.bin",
#   backingpath = bp,
#   descriptorfile = "X.desc")
# 
# # copy the data
# X_bm[,] <- matrix.ctrls
# flush(X_bm)
# 
# rm(matrix.ctrls)
# rm(X_bm)
# gc()

```

Stores data on disk and maps portions into memory as needed - MUCH more efficient
```{r}
# icloud
#bp <- "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/data/processed/bigdata"
# local
bp <- "/Users/ryan_summers/GitHub/Teddy-DNAm-Analysis/data/processed/bigdata"

matrix.clean <- attach.big.matrix(file.path(bp, "X.desc"))
dim(matrix.clean)

# probe_ids <- rownames(matrix.ctrls)
# sample_ids <- colnames(matrix.ctrls)
# 
# saveRDS(probe_ids, file.path(bp, "probe_ids.rds"))
# saveRDS(sample_ids, file.path(bp, "sample_ids.rds"))

probe_ids <- readRDS(file.path(bp, "probe_ids.rds"))
sample_ids <- readRDS(file.path(bp, "sample_ids.rds"))

```


## Exploratory Data Analysis (EDA)

```{r}
# number of patients
cat("Number of patients:",
  length(unique(pheno_complete$maskid)),"\n\n")

# number of sample wells
cat("Number of sample Wells:",
  length(unique(pheno_complete$Sample_Well)),"\n\n")

# number of sample plates
cat("Number of sample Plates:",
  length(unique(pheno_complete$Sample_Plate)),"\n\n")

# number of slides
cat("Number of slides:",
  length(unique(pheno_complete$Slide)),"\n\n")

# arrays on slide - 8
unique(pheno_complete$Array)
```

Number of CpGs on Sex chromosomes
```{r}
cat("Number of CpGs on X chromosome:",
  length(intersect(rownames(anno)[anno$chr %in% c("chrX")], probe_ids)),"\n\n")

cat("Number of CpGs on Y chromosome:",
  length(intersect(rownames(anno)[anno$chr %in% c("chrY")], probe_ids)),"\n\n")
```


```{r}
# number of patients per clinical site ≈73.5% europe
cc_distn <- pheno_complete %>% 
  group_by(cc) %>% 
  summarize(counts = n()) %>%
  mutate(percentage = round(counts / sum(counts) * 100, 2))

cc_distn %>% arrange(desc(counts))

```

### Missingness

```{r}
# missing value percentages
pheno_complete %>% select(persist_conf_ab) %>% ungroup() %>% miss_var_summary()

```

### Subjects & Obs per Age Bracket

```{r}
pheno_complete %>%
  mutate(
    age_group = cut(
      age_yrs,
      breaks = c(0, 1, 2, 3, 4, 5, 6, 7),
      include.lowest = TRUE
      )) %>%
  group_by(age_group) %>%
  summarise(
    n_subj = n_distinct(maskid),
    n_obs = n(),
    .groups = "drop") %>%
  mutate(
    subj_pct = round(n_subj / sum(n_subj) * 100, 1),
    obs_pct  = round(n_obs / sum(n_obs) * 100, 1),
    Subjects = paste0(n_subj, " (", subj_pct, "%)"),
    Observations = paste0(n_obs, " (", obs_pct, "%)")) %>%
  select(age_group, Subjects, Observations) %>%
  gt() %>%
  tab_header(
    title = md("**Age Distribution**"),
    subtitle = "Unique subjects and total observations by age group") %>%
  cols_label(age_group = "Age group") %>%
  tab_footnote(
    footnote = html("[0,1] = 0 ≤ age ≤ 1<br>(1,2] = 1 < age ≤ 2"),
    locations = cells_column_labels(columns = age_group)) %>%
  cols_align(align = "center", columns = everything()) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(13),
    column_labels.font.weight = "bold",
    table.border.top.width = px(2),
    table.border.bottom.width = px(2),
    data_row.padding = px(6))
```

[0,1] = 0 ≤ age ≤ 1
(1,2] = 1 < age ≤ 2

```{r}
# observation per subject and case/control status
pheno_complete %>%
  group_by(maskid, CaseControl) %>%
  summarise(n_obs = n(), .groups = "drop") %>% 
# group stats by case/control
  group_by(CaseControl) %>%
  summarise(
    n_subj = n(),
    mean_obs = round(mean(n_obs),1),
    median_obs = median(n_obs),
    min_obs = min(n_obs),
    max_obs = max(n_obs))
```


```{r}
# investigating class differences for clinical site
table(pheno_complete$gender, pheno_complete$cc, pheno_complete$CaseControl)
```

Check how many subjects only have one observation, which can be ill-posed for
random intercept/slope models

```{r}
# only one subject with one observation
pheno_complete$maskid <- factor(pheno_complete$maskid)
mean(table(pheno_complete$maskid) == 1) # prop
sum(table(pheno_complete$maskid) == 1) # count
```


```{r}
pheno_index <- pheno_complete %>%
  mutate(
    age_at_diag = case_endptage / 365.25,
    time_to_index = abs(age_yrs - age_at_diag)) %>%
  group_by(maskid) %>%
  slice_min(time_to_index, with_ties = FALSE) %>%
  ungroup()

pheno_longitudinal <- pheno_complete %>%
  group_by(maskid, CaseControl) %>%
  summarise(
    n_visits = n(),
    min_age = min(age_yrs, na.rm = TRUE),
    max_age = max(age_yrs, na.rm = TRUE),
    followup_years = max_age - min_age,
    .groups = "drop")

pheno_subject <- pheno_index %>%
  left_join(pheno_longitudinal, by = "maskid")


pheno_subject$dr34 <- factor(pheno_subject$dr34)

pheno_subject <- pheno_subject %>%
  mutate(
    dr34 = case_when(
      is.na(dr34) ~ NA_character_,
      dr34 == 0 ~ "No",
      dr34 == 1 ~ "Yes",
      TRUE ~ NA_character_) |> factor(levels = c("No", "Yes")))

```

### Table 1 Study Characterisitcs

```{r}
tbl1 <- pheno_subject %>%
  mutate(age_at_diag = case_endptage/365.25,
         race = factor(
           na_if(race_ethnicity, 5),
           levels = c(1, 2, 3, 4),
           labels = c("Hispanic (all races)", "NHW", "AA", "Other")),
         fdr_3level = factor(fdr_3level, levels = c("none", "dadsib", "mom"))) %>% 

  tbl_summary(
    by = CaseControl.y,
    include = c(age_at_diag, n_visits, followup_years, dr34, fdr_3level,
                gender, race, cc),
    type = list(
      all_continuous() ~ "continuous2",
      c(dr34, fdr_3level, gender, race, cc) ~ "categorical"), #
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    
    # change variable names
    # label = list(
    #   n_visits ~ "Visits per subject",
    #   age_at_diag  ~ "Age at Diagnosis",
    #   followup_years ~ "Follow-up duration (yrs)"
    # )) %>% 
    label = list(
      gender ~ "Gender",
      race ~ "Race/Ethnicity",
      dr34 ~ "HLA-DR3/DR4",
      fdr_3level ~ "First-Degree Relative",
      age_at_diag  ~ "Age at Diagnosis",
      cc ~ "Clinical Site",
      n_visits ~ "Visits per subject",
      followup_years ~ "Follow-up duration (yrs)"
      )) %>%

  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  add_n() %>% 
  # modify_spanning_header(
  #   c("stat_1", "stat_2") ~ "****") %>%
  modify_footnote_header(
    footnote = "Cases and controls were matched on age, sex, and clinical site 
    at the index timepoint; therefore, these characteristics are presented for
    the cohort overall rather than contrasted by case-control status",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  tab_header(
    title = "Table 1. Characteristics of the TEDDY Study Cohort") %>% 
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    table.width = pct(75),
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))


tbl1
gt::gtsave(tbl1, "/Users/ryan_summers/Desktop/table1.html")
```


Number of observations for each participant
```{r}
# determine number of subjects that would give "stable" estimates
subj <- pheno_complete %>%
  group_by(maskid) %>%
  summarise(
    n_obs = n(),
    n_age_unique = n_distinct(sample_agedys), 
    .groups = "drop")

# cumulative percentages (e.g. % more than 3 measurements=70.1%)
subj %>%
  reframe(
    Obs = 1:max(n_obs),
    `%` = sapply(Obs, function(k) {
      mean(n_obs >= k) * 100
      })) %>% 
  slice(1:10) %>% 
  gt() %>%
  tab_header(title = "% of Subjects With ≥ k Observations") %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title(groups = "title")) %>%
tab_options(column_labels.font.weight = "bold") %>% 
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(columns = `%`,decimals = 1) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "center",
    table.width = pct(75),
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

```

```{r}
# Confirm sample ID matching across pheno and M-matrix
sid <- as.character(pheno_complete$rgName)
mean(sid %in% colnames(matrix.filt)) # should be 1
sum(!(sid %in% colnames(matrix.filt))) # should be 0

```

Sanity check: confirm 1:1 sets
* is there a match for each case indicator? Other words, a case & control for each indicator

```{r}
check_sets <- pheno_complete %>%
  dplyr::distinct(case_ind, maskid, outcome, case_endptage) %>%
  group_by(case_ind) %>%
  summarise(
    n_ids = n_distinct(maskid),
    n_cases = sum(outcome == 1),
    n_ctrls = sum(outcome == 0),
    endpt_unique = n_distinct(case_endptage),
    .groups = "drop")

table(check_sets$n_ids) # checks number of IDs per Case Indicator

# identifies case indicators that do not have a match
bad_sets <- check_sets %>% 
  filter(n_ids != 2 | n_cases != 1 | endpt_unique != 1)
bad_sets

pheno_complete %>%
  group_by(maskid, CaseControl) %>%
  summarise(
    n_obs = n(),
    .groups = "drop") %>%
  group_by(CaseControl) %>%
  summarise(
    n_subj = n(),
    mean_obs = mean(n_obs),
    median_obs = median(n_obs),
    min_obs = min(n_obs),
    max_obs = max(n_obs))

# look at some matched samples
pheno_complete[
  pheno_complete$maskid==997784, 
  c("maskid", "gender", "CaseControl", "cc", "sample_agedys", "sample_visit")] %>% 
  arrange(sample_visit)
```

### CpG Reports

```{r}
# generate a CpG report
rep <- cpg_report(
  "cg26507939", 
  matrix.clean,
  pheno_complete,
  anno,
  DNAm_values = "M",
  outlier_method = "mad",
  z_thresh = 3)

# extract CpG values for each subject along w/ pheno data
rep$data

# annotation
rep$anno

# get outliers
rep$outliers %>% arrange(desc(abs(outlier_score)))
```

```{r}
sample_outlier_counts <- outliers %>%
  dplyr::count(sample)

outliers %>% 
  group_by(CpG) %>% 
  summarize(counts = n()) %>% 
  arrange(desc(counts))
```

1.25% (~13.9M) of total observations were flagged as outliers

### CpG Variability (Outlier Detection)

Outlier detection/removal for modeling 
```{r}
# MAD per CpG - **ignores trajectory and age ordering**
mad_vals <- rowMads(matrix.filt, constant = 1.4826, na.rm = TRUE)

# mad z-scores
zMAD <- sweep(matrix.filt, 1, med, "-")
zMAD <- sweep(zMAD, 1, mad_vals, "/") # matrix of MAD scores

# outlier matrix
outlier_mat <- abs(zMAD) >= 3 #??
```



$\text{MAD} =\text{median}\left(|x_i - \text{median}(x)|\right)$
```{r}
cpg_mad <- numeric(nrow(matrix.clean))
# chunk size is 1000 - test 5000 for speed
for (start in seq(1, nrow(matrix.clean), by = 1000)) {
  end <- min(start + 1000 - 1, nrow(matrix.clean))
  
  # pull only a chunk (1000 CpGs) into memory at a time
  mat_chunk <- matrix.clean[start:end, ]
  
  # row-wise MAD
  cpg_mad[start:end] <- matrixStats::rowMads(as.matrix(mat_chunk), na.rm = T)
}

names(cpg_mad) <- probe_ids
```

```{r}
summary(cpg_mad)
# test a threshold
threshold <- quantile(cpg_mad, 0.10, na.rm = TRUE)

low_var_cpgs <- names(sort(cpg_mad[cpg_mad <= threshold],decreasing = T))

length(low_var_cpgs)
head(low_var_cpgs)
```

Visit/age-window-specific IQR filtering
```{r}
pheno_complete <- pheno_complete %>%
  mutate(
    age_bin = cut(
      age_yrs,breaks = c(0, 0.5, 1, 2, 3, 4, 6.5), include.lowest = T, right = F,
      labels = c("0-6mo", "6-12mo", "1-2yr", "2-3yr", "3-4yr", "4-6yr")
    ))

# initialize a new matrix
matrix.iqr.clean <- matrix.ctrls
# outlier matrix
outlier_mat_iqr <- 
  matrix(FALSE, nrow = nrow(matrix.ctrls), ncol = ncol(matrix.ctrls))
rownames(outlier_mat_iqr) <- rownames(matrix.ctrls)
colnames(outlier_mat_iqr) <- colnames(matrix.ctrls)

k <- 5   # try 3, 5, 7 as sensitivity - 

bins <- unique(na.omit(pheno_complete$age_bin))

plan(multisession, workers = 12)  
options(future.globals.maxSize = 5 * 1024^3)

min_iqr <- 0.1  # on the m-value scale (play around with this)

outlier_list <- future_lapply(bins, function(bin) {

  samples_bin <- pheno_complete$rgName[
    pheno_complete$age_bin == bin &
      !is.na(pheno_complete$age_bin)]

  samples_bin <- intersect(samples_bin, colnames(matrix.ctrls))

  if (length(samples_bin) < 10) {
    return(NULL)
  }

  X <- matrix.ctrls[, samples_bin, drop = FALSE]

  q1 <- matrixStats::rowQuantiles(X, probs = 0.25, na.rm = TRUE)
  q3 <- matrixStats::rowQuantiles(X, probs = 0.75, na.rm = TRUE)
  iqr <- q3 - q1

  valid_iqr <- is.finite(iqr) & iqr > min_iqr

  out_bin <- matrix(F, nrow = nrow(X), ncol = ncol(X), dimnames = dimnames(X))

  if (any(valid_iqr)) {
    lower <- q1[valid_iqr] - k * iqr[valid_iqr]
    upper <- q3[valid_iqr] + k * iqr[valid_iqr]

    X_valid <- X[valid_iqr, , drop = FALSE]

    out_bin[valid_iqr, ] <-
      sweep(X_valid, 1, lower, "<") |
      sweep(X_valid, 1, upper, ">")
  }

  out_bin[is.na(out_bin)] <- FALSE

  list(
    samples = samples_bin,
    out_bin = out_bin,
    n_valid_iqr = sum(valid_iqr),
    n_skipped_low_iqr = sum(!valid_iqr, na.rm = TRUE)
  )
},
future.seed = TRUE)

plan(sequential)

# plug results into outlier matrix
for (res in outlier_list) {
  if (is.null(res)) next
  outlier_mat_iqr[, res$samples] <- res$out_bin
  }

# label outliers as NA in matrix
matrix.iqr.clean[outlier_mat_iqr] <- NA
qsave(matrix.iqr.clean, "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/matrix.iqr.clean.qs", nthreads = 12)

```

```{r}
# save outlier records
idx <- which(outlier_mat_iqr, arr.ind = T)

iqr_outliers <- data.frame(
  CpG = rownames(matrix.ctrls)[idx[, 1]],
  sample = colnames(matrix.ctrls)[idx[, 2]],
  value = matrix.ctrls[idx],
  age_yrs = pheno_complete$age_yrs[
    match(colnames(matrix.ctrls)[idx[, 2]], pheno_complete$rgName)],
  age_bin = pheno_complete$age_bin[
    match(colnames(matrix.ctrls)[idx[, 2]], pheno_complete$rgName)])

matrix.iqr.clean <- matrix.ctrls[iqr_outliers$sample]

# iqr_outliers <- qload(iqr_outliers, "/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/iqr_outliers.qs", nthreads = 12)

```

```{r}
length(unique(iqr_outliers$CpG))
length(unique(iqr_outliers$sample))
table(iqr_outliers$age_bin)
mean(table(iqr_outliers$sample))

# outliers pers sample
table(iqr_outliers$age_bin) /table(pheno_complete$age_bin)
```


```{r}

for (i in sample(iqr_outliers$CpG, 50)){
  p <- getSpagPlot_big(
    CpG = i, # cg20931951 cg04422591 cg01593740 cg12109275
    mvals_bm = matrix.clean,
    value_type = "beta", #beta or m
    age_unit = "years",
    y_limits = c(0,1),
    pheno = pheno_complete,
    anno = anno,
    probe_ids = probe_ids,
    sample_ids = sample_ids)
  print(p)
  }
```


```{r}
getSpagPlot_big(
  CpG = "cg19565594",
  mvals_bm = matrix.clean,
  pheno = pheno_complete,
  anno = anno,
  probe_ids = rownames(matrix.ctrls),
  sample_ids = colnames(matrix.ctrls),
  value_type = "beta",
  age_unit = "years",
  highlight_sample = 
    c("206154070036_R01C01", "206154260058_R02C01", "206154260009_R01C01")
  )

iqr_outliers %>%
  filter(CpG == "cg19565594") %>%
  select(CpG, sample, value, age_bin)
```


## Sample Modeling Runs

```{r}
# example CpG
probe <- "cg26928153"
i <- match(probe, probe_ids) # find match

if (is.na(i)) stop("Probe not found: ", probe)

pheno <- pheno_complete

# pull methylation values for one CpG
cpg <- as.numeric(matrix.clean[i, ])
names(cpg) <- sample_ids

# attach CpG values to phenotype by sample ID
pheno$CpG <- cpg[as.character(pheno$rgName)]

# keep only variables needed for the model
vars_needed <- c(
  "CpG", "maskid", "age_yrs",
  "gender", "cc", "CaseControl",
  "new_Bcell", "new_CD4T", "new_CD8T", "new_Mono", "new_NK")

dat <- pheno[complete.cases(pheno[, vars_needed]), vars_needed, drop = FALSE]

# fit model
m1 <- lmer(
  CpG ~ ns(age_yrs, df = 3) + gender + cc +
    new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK +
    (1 | maskid),
  data = dat,
  REML = T)

m2 <- lmer(
  CpG ~ bs(age_yrs, df = 3) + gender + cc +
    new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK +
    (1 | maskid),
  data = dat,
  REML = T)

# # cell composition is confounding the age trend and dominates the signal
# m3 <- lmer(
#   CpG ~ age_yrs + gender + cc + 
#     new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK +
#     (1 | maskid),
#   data = dat,
#   REML = T)


summary(m1)
summary(m2)
#summary(m3)
anova(m1,m2)

```

```{r}
pred_res <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_predictions_ctrls.qs", nthreads = 12)
```



Wald statistic calculation
```{r}
# variance-covariance matrix for spline terms
vc <- vcov(m1)[
  c("ns(age_yrs, df = 3)1","ns(age_yrs, df = 3)2","ns(age_yrs, df = 3)3"), 
  c("ns(age_yrs, df = 3)1","ns(age_yrs, df = 3)2","ns(age_yrs, df = 3)3")]

# upper-triangle values needed for joint Wald chi-sq test
vc11 <- vc[1,1]; vc22 <- vc[2,2]; vc33 <- vc[3,3]
vc12 <- vc[1,2]; vc13 <- vc[1,3]; vc23 <- vc[2,3]
```

```{r}
fe <- fixef(m1)
betas <- c(
  fe[["ns(age_yrs, df = 3)1"]], 
  fe[["ns(age_yrs, df = 3)2"]], 
  fe[["ns(age_yrs, df = 3)3"]])

V <- matrix(c(vc11, vc12, vc13,
              vc12, vc22, vc23,
              vc13, vc23, vc33), nrow = 3)

wald_stat <- t(betas) %*% solve(V) %*% betas
pchisq(wald_stat, df = 3, lower.tail = F)

# use contest() ftn instead*****
```

```{r}
coef_names <- names(fixef(m1))

ns_names <- paste0("ns(age_yrs, df = 3)", 1:3)
ns_idx <- match(ns_names, coef_names)

L <- matrix(
  0,
  nrow = 3,
  ncol = length(coef_names),
  dimnames = list(ns_names, coef_names))

L[cbind(1:3, ns_idx)] <- 1

lmerTest::contestMD(m1, L=L, rhs = 0,ddf = "Satterthwaite")
```
Predicted methylation at age (a) toy example
```{r}

# Ages at which to evaluate the model
age_grid <- as.numeric(
  quantile(
    dat$age_yrs, 
    probs = seq(0, 1, length.out = 12),
    na.rm = T))

beta <- fixef(m1)
Vbeta <- as.matrix(vcov(m1))

random_intercept_var <- as.numeric(VarCorr(m1)$maskid[1, 1])
residual_var <- sigma(m1)^2

# These terms retain the spline knots used when m1 was fitted
fixed_terms <- delete.response(terms(m1, fixed.only = T))

# Retain factor levels and contrasts from the fitted model
pred_list <- lapply(age_grid, function(a) {

  dat$age_yrs <- a

  # Build the model frame using the terms stored in m1
  mf_new <- model.frame(
    fixed_terms,
    data = dat,
    xlev = .getXlevels(terms(m1, fixed.only = T), model.frame(m1)),
    na.action = na.pass)

  # This now reuses the original spline knots
  X <- model.matrix(
    fixed_terms,
    data = mf_new,
    contrasts.arg = attr(getME(m1, "X"), "contrasts"))

  X <- X[, names(beta), drop = F]

  # Average across the observed covariate distribution
  xbar <- colMeans(X)
  estimate <- as.numeric(xbar %*% beta)
  fixed_var <- as.numeric(xbar %*% Vbeta %*% xbar)
  conf_se <- sqrt(fixed_var)

  pred_se <- sqrt(fixed_var + random_intercept_var + residual_var)

  data.frame(
    age_yrs = a,
    CpG_est = estimate,

    conf_se = conf_se,
    cLCL = estimate - 1.96 * conf_se,
    cUCL = estimate + 1.96 * conf_se,

    pred_se = pred_se,
    pLL = estimate - 1.96 * pred_se,
    pUL = estimate + 1.96 * pred_se
  )
})

pred_dat <- do.call(rbind, pred_list)

head(pred_dat)
```

Results from local spline model runs
```{r}
comparisons_tbl <- qread(
  "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/comparison_tbl.qs",
  nthreads = 12)
```

```{r}
model_preferences <- comparisons_tbl %>%
  filter(spline_df %in% c(3, 4, 5), is.finite(AIC), is.finite(BIC)) %>%
  dplyr::select(CpG, spline_df, AIC, BIC) %>%
  pivot_wider(
    names_from = spline_df,
    values_from = c(AIC, BIC),
    names_glue = "{.value}_df{spline_df}"
  ) %>%
  # Require all three models
  dplyr::filter(
    complete.cases(
      AIC_df3,
      AIC_df4,
      AIC_df5,
      BIC_df3,
      BIC_df4,
      BIC_df5
    )
  ) %>%
  dplyr::mutate(
    ##########################################################
    # Pairwise differences
    #
    # Positive values mean the first/lower-df model is better.
    # Example:
    # AIC_df4_minus_df3 > 0 means df3 has lower AIC.
    ##########################################################

    AIC_df4_minus_df3 = AIC_df4 - AIC_df3,
    AIC_df5_minus_df4 = AIC_df5 - AIC_df4,
    AIC_df5_minus_df3 = AIC_df5 - AIC_df3,

    BIC_df4_minus_df3 = BIC_df4 - BIC_df3,
    BIC_df5_minus_df4 = BIC_df5 - BIC_df4,
    BIC_df5_minus_df3 = BIC_df5 - BIC_df3,

    ##########################################################
    # Minimum AIC and BIC for each CpG
    ##########################################################

    minimum_AIC = pmin(AIC_df3,AIC_df4,AIC_df5),

    minimum_BIC = pmin(BIC_df3,BIC_df4,BIC_df5),

    ##########################################################
    # Difference from the best model
    ##########################################################

    delta_AIC_df3 = AIC_df3 - minimum_AIC,
    delta_AIC_df4 = AIC_df4 - minimum_AIC,
    delta_AIC_df5 = AIC_df5 - minimum_AIC,

    delta_BIC_df3 = BIC_df3 - minimum_BIC,
    delta_BIC_df4 = BIC_df4 - minimum_BIC,
    delta_BIC_df5 = BIC_df5 - minimum_BIC,

    ##########################################################
    # Number of models tied for the minimum
    ##########################################################

    n_AIC_minima = (
      abs(AIC_df3 - minimum_AIC) < 1e-8) + 
      (abs(AIC_df4 - minimum_AIC) < 1e-8) + 
      (abs(AIC_df5 - minimum_AIC) < 1e-8),

    n_BIC_minima = (
      abs(BIC_df3 - minimum_BIC) < 1e-8) + 
      (abs(BIC_df4 - minimum_BIC) < 1e-8) + 
      (abs(BIC_df5 - minimum_BIC) < 1e-8),

    ##########################################################
    # Strictly preferred model
    ##########################################################

    preferred_AIC = dplyr::case_when(
      n_AIC_minima > 1 ~ "tie",
      abs(AIC_df3 - minimum_AIC) < 1e-8 ~ "df3",
      abs(AIC_df4 - minimum_AIC) < 1e-8 ~ "df4",
      abs(AIC_df5 - minimum_AIC) < 1e-8 ~ "df5",
      TRUE ~ NA_character_
    ),

    preferred_BIC = dplyr::case_when(
      n_BIC_minima > 1 ~ "tie",
      abs(BIC_df3 - minimum_BIC) < 1e-8 ~ "df3",
      abs(BIC_df4 - minimum_BIC) < 1e-8 ~ "df4",
      abs(BIC_df5 - minimum_BIC) < 1e-8 ~ "df5",
      TRUE ~ NA_character_
    )
  )
```

```{r}
preference_summary <- bind_rows(
  model_preferences %>%
    dplyr::count(preferred_AIC, name = "n_CpGs") %>%
    transmute(
      criterion = "AIC",
      preferred_model = preferred_AIC,
      n_CpGs),

  model_preferences %>%
    dplyr::count(preferred_BIC, name = "n_CpGs") %>%
    transmute(
      criterion = "BIC",
      preferred_model = preferred_BIC,
      n_CpGs)) %>%
  group_by(criterion) %>%
  mutate(percent = 100 * n_CpGs / sum(n_CpGs)) %>%
  ungroup()

preference_summary
```

Inspection of saved models from toy run
```{r}
# # load in models
# toy_results <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/toy_spline_model_objects.qs", nthreads=12)

m_cg_df3 <- toy_results[["cg26928153"]]$models[["ns_df3"]]
m_cg_df4 <- toy_results[["cg26928153"]]$models[["ns_df4"]]

summary(m_cg_df3)
summary(m_cg_df4)

AIC(m_cg_df3, m_cg_df4)
BIC(m_cg_df3, m_cg_df4)

isSingular(m_cg_df3)
isSingular(m_cg_df4)

plot(m_cg_df3)
qqnorm(residuals(m_cg_df3))
qqline(residuals(m_cg_df3))
```


Spline boundary knots
```{r}
ns_basis <- ns(pheno_complete$age_yrs[pheno_complete$rgName %in% sample_ids], 
               df = 3) 
attr(ns_basis, "knots")
attr(ns_basis, "Boundary.knots")
```


```{r}
getSpagPlot_big(
  CpG = "cg26928153",
  mvals_bm = matrix.clean,
  value_type = "beta",
  #y_limits = c(0.85, 1),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  #group_col = "CaseControl",
  #facet_by_group = T,
  legend = "none")

```



```{r}
qqnorm(resid(m1))
qqline(resid(m1))
```


```{r}
# diagnose singular fit
re <- ranef(m3)$maskid

summary(re$age_yrs)
hist(re$age_yrs)

plot(fitted(m3), resid(m3))
abline(h = 0, col = "red")

vc <- VarCorr(m3)
G <- as.matrix(vc$maskid)
eigen(G) # $values[1] is int and $values[2] is slope

dat %>%
  group_by(maskid) %>%
  summarise(
    n_visits = n(),
    age_range = max(age_yrs) - min(age_yrs)) %>%
  summary()
```



```{r}
plot_df <- data.frame(
  age = rep(ages, 3),
  sample_agedys = rep(ages * 365.25, 3),
  pred = c(pred_lm, pred_lmer_int, pred_lmer_slope),
  model = rep(c("LM", "LMER_int", "LMER_slope"), each = length(ages)))

plot_df <- plot_df %>%
  mutate(pred_beta = m2beta(pred))

ggplot(plot_df, aes(age, m2beta(pred), color = model)) +
  geom_line(linewidth =0.8) +
    geom_point()

getSpagPlot_big(
  CpG = "cg19159092",
  mvals_bm = matrix.clean,
  value_type = "beta", #beta or m
  y_limits = c(0,1),
  pheno = pheno,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  pred_df = plot_df)
```


### Plotting Spline Trajectories

```{r}
cpg_id <- "cg12025418"
plot_dt <- melt(
  lmeR_spline_int_res[CpG == cpg_id],
  id.vars = "CpG",
  measure.vars = pred_cols,
  variable.name = "age_col",
  value.name = "pred")

plot_dt[, age := as.numeric(sub("^pred_age_", "", age_col))]

ggplot(plot_dt, aes(x = age, y = pred)) +
  geom_line(size = 1) +
  geom_point() +
  theme_classic() +
  labs(
    title = paste("Trajectory for", cpg_id),
    x = "Age (years)",
    y = "Predicted methylation")
```

```{r}
getSpagPlot_big(
  CpG = "cg20389709",
  mvals_bm = matrix.clean,
  value_type = "beta", #beta or m
  y_limits = c(0,1),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids)
```


### Plot Spline Basis Functions

```{r}
age_seq <- seq(
  min(dat$age_yrs), max(dat$age_yrs, na.rm = TRUE), 
  length.out = 200)

B <- ns(age_seq, df = 3)

basis_df <- data.frame(
  age_yrs = age_seq,
  B1 = B[, 1],
  B2 = B[, 2],
  B3 = B[, 3])

basis_long <- pivot_longer(
  basis_df,
  cols = c(B1, B2, B3),
  names_to = "basis",
  values_to = "value")

# not interpretable individually - meaningful when multiplied by β and summed
ggplot(basis_long, aes(x = age_yrs, y = value, color = basis)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    x = "Age (years)",
    y = "Spline basis value",
    title = "Natural spline basis functions: ns(age_yrs, df = 3)")
```

We evaluated potential effect modification of age by sex, clinical site, and 
case/control status in a subset of CpGs. Across these analyses, inclusion of 
interaction terms did not improve model fit (based on likelihood ratio tests 
and AIC), suggesting that age-related methylation trajectories were broadly 
consistent across strata. Therefore, we summarized CpG trajectories using 
marginal predictions averaged over the observed covariate distribution.

To avoid perfect collinearity among estimated cell proportions, one cell type 
(neutrophils) was omitted from the model and served as the reference component. 
Thus, coefficients for included cell proportions are interpreted relative to 
the omitted neutrophil fraction, while age trajectories represent methylation 
patterns adjusted for measured cell-composition differences


"Model failed to converge with max|grad| = 0.00399344 (tol = 0.002, component 1)":

That warning means lmer returned estimates, but the optimizer stopped with a gradient that’s a bit larger than the default tolerance. In plain terms:

* the algorithm got close to the optimum, but not close enough to declare full convergence.

So the results aren’t automatically “wrong,” but you should treat that CpG’s fit as potentially unreliable (especially SEs / variance components).


Eigenvalue: 
0.0345 - Substantial random-effects variability
5×10⁻⁹- Essentially zero variability

Interpretation:
There is strong subject-to-subject variability in one direction (intercept), 
and almost none in the other (slope-related direction).

* Eigenvalues > 0 -> matrix is positive definite (good)
* Very small eigenvalues -> one direction has almost no variance
*	≈ 0 eigenvalue -> model is near-singular

Main takeaways:

Group 1 (~50%)
	* mostly linear
	* spline ≈ linear

Group 2 (~35%) (Δ in AIC > 2)
	* moderately nonlinear
	* spline changes curvature but not drastically

Group 3 (~15–20%) (Δ AIC > 10)
	* strongly nonlinear
	* spline fundamentally changes trajectory


* The REML likelihood depends on the fixed‐effects (FE) design matrix
* REML integrates out fixed effects
* If you compare models with different fixed effects, the REML likelihoods are
not comparable

## Model Results  

### Model Formulation for linear mixed model (LMM)

* 10 fixed effects
* 3 random effects (intercept, slope, cov(int,slope))
* 1 residual variance $\sigma^2$ (within)
* k = 14

Model:
$$
Y_i
= \beta_0

+ \beta_{\text{age}}(\text{age}_i)
+ \beta_{\text{gender}}(\text{gender}_i)
+ \beta_{\text{cc}}(\text{CC}_i)
+ \beta_{\text{CaseControl}}(\text{CaseControl}_i)
+ \beta_{\text{B}}(\text{Bcell}_i) \\
+ \beta_{\text{CD4}}(\text{CD4T}_i)
+ \beta_{\text{CD8}}(\text{CD8T}_i)
+ \beta_{\text{Mono}}(\text{Mono}_i)
+ \beta_{\text{NK}}(\text{NK}_i) \\
+ b_{0}
+ b_{1}(\text{age}_i)
+ \varepsilon_i
$$

$$
\mathbf y_i \sim N(X_i\beta,\ V_i) \\
\mathbf y_i = X_i\beta + Z_i b_i + \epsilon_i
$$


log-likelihood:
$$
\ell =
-\frac{1}{2}
\left[
\log|V|

+ (y-X\beta)^T V^{-1}(y-X\beta)
+ n \log(2\pi)
  \right]
$$

$$
\hat\beta = (X^\top V^{-1} X)^{-1}X^\top V^{-1} y
$$
Subject-level
$$
V_i = Var(Y_i) = Z_iGZ^T_i + R_i
$$
	*	$V_i$ = total variability of $Y_i$
	*	from random effects $Z_i G Z_i^\top$
	*	plus noise $R_i$

$$
\hat{b}_i = E(b_i \mid Y_i) = G Z_i^\top V_i^{-1}(Y_i - X_i \hat{\beta})
$$
This gives the best prediction of the subject-specific random effects $b_i$, given the observed data $Y_i$.
  * So b_i captures how individual $i$ deviates from the population trend.

$V_i^{-1}$:
	*	weights the residuals
	*	gives less weight to noisy observations
	*	accounts for correlation in repeated measures

$$
\hat{b}_i = \text{(variance)} \times \text{(design)}^\top \times \text{(precision)} \times \text{(residual)}
$$


Random effects Var-Cov Matrix
$$
G=
\begin{pmatrix}
\sigma^2_{intercept} & \sigma_{int,slope}\\
\sigma_{int,slope} & \sigma^2_{slope}
\end{pmatrix}
$$

$$
x^T G x = \mathrm{Var}(x^T b) \ge 0
$$
* Strictly positive unless there is perfect linear dependence. I.e symmetric positive-definite (SPD)
* In linear mixed models, covariance matrices are symmetric positive-definite (or semidefinite), and the marginal covariance $V = Z G Z^T + \sigma^2 I$ is SPD when residual variance is positive.
    * $G$ describes variability of random effects
    * $Z G Z^T$ induces correlation between observations
    * $\sigma^2 I$ adds independent noise
    * total covariance must represent a real variance -> SPD
    * if SPD $G = LL^T$ - where $L$ is lower triangular : 
        $\text{det}(G) = (\prod L_{ii})^2 = \prod_i \lambda_i$
      * $\text{det}(G)$ represents total variability of the random effects
      * large determinant -> random effects spread widely
	    * small determinant -> random effects concentrated
	    * determinant = 0 -> ellipsoid collapses -> singular covariance

Interpretation for CpG cg26928153:
***Must know the baseline methylation value.***
$$
\frac{d\beta}{dM} = \beta(1-\beta) \text{ln}(2) \\
\frac{d\beta}{dage} = \beta(1-\beta) \text{ln}(2) \times \frac{dM}{dage} \text{(our model estimate)}
$$

$$
M(t)=\text{fixed_intercept}+\text{fixed_slope}\cdot t \\

\beta(t)=m2beta(M(t)) \\

\frac{d\beta}{d age}(t)=\beta(t)(1-\beta(t))\ln(2)\cdot \text{fixed_slope}
$$

Say baseline β:
* β=0.5: β(1-β) = 0.25 * ln(2) * β1 = 0.25 x ln(2) x 0.000591 = 0.000102
* β=0.2: β(1-β) = 0.16 * ln(2) * β1 = 0.16 x ln(2) x 0.000591 = 0.000065
* β=0.9: β(1-β) = 0.09 * ln(2) * β1 = 0.09 x ln(2) x 0.000591 = 0.000037

**CpGs near 0.5 -> biggest β change**
**CpGs near 0 or 1 -> tiny β change**

The fixed slope of 0.00059 in M-values corresponds to an approximate β-value increase of 0.00004–0.00010 per year depending on baseline methylation level, indicating a very gradual increase in methylation

In model form:
* For every 1 year increase in age, the log2-odds methylation increases by 0.000591
* log2OR = $2^{0.000591} = 1.00041$: A 1yr increase in age is associated with
  approximately a 0.041% increase in methylation odds per year 
  (or a 1.00041-fold change in methylation odds)

### Spline Interpretation 

We predicted methylation values at selected ages to reconstruct each CpG’s 
developmental trajectory.

The spline model estimates a smooth methylation curve using piecewise cubic
polynomials. The predicted  values are points sampled from that curve at 
specific ages, and connecting those points reconstructs the estimated trajectory
across childhood.
* basis functions (ns_est) are the building blocks of the curve
* coefficients determine how much each "block" contributes
* each basis function (B) is a cubic polynomial-based basis function that 
  together form a piecewise cubic curve
    * each B is a curved template and coefficients determine how strong the 
      template influences the final curve
    * large coefficient has stronger pull on shape, whereas a small coefficient
      provides weaker contribution

Analogy:

Think of the spline basis functions as flexible pieces of curved wire and the 
coefficients determine how strongly each wire bends the final trajectory.

The final curve is the weighted combination:
$$
\hat{y}_{ij}(age)
=
\hat{\beta}_0
+
\hat{\beta}_1 B_1(age)
+
\hat{\beta}_2 B_2(age)
+
\hat{\beta}_3 B_3(age)
+
\mathbf{x}_{ij}^T\hat{\gamma}
$$

where:
* $B_1(age), B_2(age), B_3(age)$
    are the spline basis functions
* $\hat{\beta}_1, \hat{\beta}_2, \hat{\beta}_3$
    are the estimated spline coefficients
* $\mathbf{x}_{ij}^T\hat{\gamma}$
    are covariate contributions
* $\hat{\beta}_0$
    is the intercept
    
```{r}
ns_knots <- splines::ns(pheno_complete$age_yrs,df = 3)
attr(ns_knots, "knots")
```


```{r}
library(splines)

x <- pheno_complete$age_yrs

# fit natural spline basis
ns_fit <- ns(x, df = 3)

knots <- attr(ns_fit, "knots")
boundary <- attr(ns_fit, "Boundary.knots")

knots
boundary
age_grid <- seq(min(x), max(x), length.out = 200)

B <- predict(ns_fit, newx = age_grid)

matplot(
  age_grid,
  B,
  type = "l",
  lty = 1,
  lwd = 2,
  xlab = "Age",
  ylab = "Spline basis value",
  main = "Natural spline basis functions"
)

abline(v = knots, lty = 2)
```



Average M-value at a specific age - calculated per CpG
$$
\widehat{m}(age)
=
\frac{1}{N}
\sum_{i=1}^{S} \sum_{j=1}^{n_i}
\hat{y}_{ij}(age)
$$
where:

* $S$ = number of subjects (209)
* $n_i$ = number of visits for subject $i$
    
We are currently in the method-development and benchmarking phase, where we are
evaluating different longitudinal modeling and clustering strategies to identify
the most biologically meaningful representation of developmental methylation 
trajectories.
  * Development and benchmarking of a longitudinal trajectory-clustering framework

In DNAm data:
	*	Blood cell proportions change strongly with age
	*	Different cell types have very different methylation profiles
Therefore, cell composition is likely explaining most of the apparent age-related 
change CpGs

### Load Results


vroom = 5.95236 secs
fread = 2.61801
read_csv = 12.89397 secs


```{r}
# spline with DF=3 - random int/slope w/ chrX
#/Library/Mobile\ Documents/com~apple~CloudDocs
lmeR_spline_results <- fread("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/lmeR_spline_results.csv") %>% select(-1)
```

```{r}
# slope models with chrX
lmeR_spline_int_full <- fread("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/lmeR_spline_int_res.csv") %>% select(-1)
```

```{r}
# spline models w/out chrX
lmeR_spline_int_slope_res <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary.qs",nthreads=12)

# controls & intercept only w/out chrX
spline_int_results <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary_ctrls.qs",nthreads=12)

# icloud
lmeR_spline_int_res <- qread("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary_inty.qs",nthreads=6)

lmeR_spline_int_slope_res <- qread("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary.qs",nthreads=6)

```

Controls only:
```{r}
# splines intercepts 
spline_int_BLUPs_full <-qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_blup_ctrls_full.qs", nthreads=12)

# spline summary results
spline_summary_BLUPs <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary_ctrls.qs",nthreads=12)
```

Controls only w/ sub-trajectories removed from GMM step:
```{r}
# splines intercepts 
spline_int_BLUPs <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_blup_ctrls.qs",nthreads=12)

# spline summary results
spline_summary_BLUPs <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_summary_ctrls.qs",nthreads=12)

# prediction stats
spline_pred_stats <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/secondary/lmeR_spline_RE_predictions_ctrls.qs",nthreads=12)
```

```{r}
spline_summary_BLUPs[CpG == "cg21960840", 
                 .(CpG, nObs, aic, is_singular)]
```

```{r}
lmeR_spline_summary[CpG == "cg06638795",
                     .SD, # SD = subset data for data.table formates
                     .SDcols = c("CpG", "nObs", "aic", "is_singular",pred_cols)]
lmeR_spline_int_res[CpG == "cg06638795",
                     .SD, # SD = subset data for data.table formates
                     .SDcols = c("CpG", "nObs", "aic", "is_singular",pred_cols)]
```

```{r}
suppressWarnings( # Warnings relate to missing values which I am aware of
for (i in cpg_diff$CpG[1:50]){
  p <- getSpagPlot_big(
  CpG = i,
  mvals_bm = matrix.clean,
  pheno = pheno_complete,
  anno = anno,
  value_type = "beta", #beta or m
  y_limits = c(0,1),
  age_unit = "years",
  probe_ids = probe_ids,
  sample_ids = sample_ids)
  print(p)
})
```


```{r}
lmeR_spline_summary %>% group_by(is_singular) %>% summarise(n = n())
lmeR_spline_int_res %>% group_by(is_singular) %>% summarise(n = n())
```


#### Contrasts Error

Identify why 1,435 CpGs produced error message "contrasts can be applied only to..."
```{r}
# example CpG
probe <- "cg26507939"
i <- match(probe, probe_ids) # find match

if (is.na(i)) stop("Probe not found: ", probe)

pheno <- pheno_complete

# pull methylation values for one CpG
cpg <- as.numeric(matrix.clean[i, ])
names(cpg) <- sample_ids

# attach CpG values to phenotype by sample ID
pheno$CpG <- cpg[as.character(pheno$rgName)]

# keep only variables needed for the model
vars_needed <- c(
  "CpG", "maskid", "age_yrs",
  "gender", "cc", "CaseControl",
  "new_Bcell", "new_CD4T", "new_CD8T", "new_Mono", "new_NK")

dat <- pheno[complete.cases(pheno[, vars_needed]), vars_needed, drop = FALSE]


table(dat$gender, useNA = "ifany") # culprit for cg26507939 - only female samples
table(dat$cc, useNA = "ifany")
table(dat$CaseControl, useNA = "ifany")
```


### Multiple Testing Adjustment

```{r}
## Should compare and contrast when removing singular slope fits for MultTest adj -if
## we decide to move along that route

# keep CpGs w/ no singular fits
lmeR_MAD_results_filt <- lmeR_MAD_results %>% 
  filter(is_singular == TRUE)

# Benjamini-Hochberg FDR 
lmeR_MAD_results_filt$q_slope <- p.adjust(lmeR_MAD_results_filt$p_slope, 
                                            method = "BH")

lmeR_MAD_results$q_slope <- p.adjust(lmeR_MAD_results$p_slope, 
                                            method = "BH")
```


```{r}
cat(
  "Non-singular fits w/ FDR (BH) adjustment and MAD outliers removed had",
  nrow(lmeR_MAD_results_filt[lmeR_MAD_results_filt$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n\n")

cat(
  "Singular fits w/ FDR (BH) adjustment and MAD outliers removed had",
  nrow(lmeR_MAD_results[lmeR_MAD_results$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n")
```

* With outliers removed, we retained 9139 more CpGs after FDR (BH) adjustment 
with a significant slope and wasn't singular

### Comparison of p-values with βeta values and AIC

Let's discover any relationships with our p-values/βetas/AIC. Looking at these 
together may reveal patterns about signal strength, practical relevance, and 
model structure.

```{r}
ggplot(lmeR_bobyqa_results, 
       aes(x = fixed_slope, y = -log10(p_slope), color = aic)) +
  geom_point(alpha = 0.25, size = 0.6) +
  scale_color_viridis_c(name = "AIC") +
  labs(x = "Fixed age slope (M-value scale)", y = "-log10(p-value)") +
  theme_classic()
```


Beta to M:
0.1	-> -3.2
0.25 -> -1.6
0.5	-> 0
0.75 ->	1.6
0.9 -> 3.2

### CpG Review

```{r}
lmeR_MAD_results[CpG %in% cpg_wrg, 
                 .(CpG, nObs, fixed_slope, is_singular, varcov.rand.eigen.2)]

lmeR_spline_results[CpG %in% cpg_wrg, 
                     .SD, # SD = subset data for data.table formates
                     .SDcols = c("CpG", "nObs", pred_cols)]

lmeR_spline_results[CpG=="cg13371836",]
```

### Spline Explanation 

We predicted methylation values at selected ages to reconstruct each CpG’s 
developmental trajectory.

The spline model estimates a smooth methylation curve using piecewise cubic
polynomials. The predicted values are points sampled from that curve at specific 
ages, and connecting those points reconstructs the estimated trajectory across
childhood.
* basis functions (ns_est) are the building blocks of the curve
* coefficients determine how much each "block" contributes
* each basis function (B) is a cubic polynomial-based basis function that 
  together form a piecewise cubic curve
    * each B is a curved template and coefficients determine how strong the 
      template influences the final curve
    * large coefficient has stronger pull on shape, whereas a small coefficient
      provides weaker contribution

Analogy:

Think of the spline basis functions as flexible pieces of curved wire and the 
coefficients determine how strongly each wire bends the final trajectory.

The final curve is the weighted combination:
$$
\hat{y}_{ij}(age)
=
\hat{\beta}_0
+
\hat{\beta}_1 B_1(age)
+
\hat{\beta}_2 B_2(age)
+
\hat{\beta}_3 B_3(age)
+
\mathbf{x}_{ij}^T\hat{\gamma}
$$
where:
* $B_1(age), B_2(age), B_3(age)$
    are the spline basis functions
* $\hat{\beta}_1, \hat{\beta}_2, \hat{\beta}_3$
    are the estimated spline coefficients
* $\mathbf{x}_{ij}^T\hat{\gamma}$
    are covariate contributions
* $\hat{\beta}_0$
    is the intercept
    
We are currently in the method-development and benchmarking phase, where we are
evaluating different longitudinal modeling and clustering strategies to identify
the most biologically meaningful representation of developmental methylation 
trajectories.
  * Development and benchmarking of a longitudinal trajectory-clustering framework

In DNAm data:
	*	Blood cell proportions change strongly with age
	*	Different cell types have very different methylation profiles
Therefore, cell composition is likely explaining most of the apparent age-related 
change CpGs

how much $R^2$ drops when removing each variable
```{r}
library(performance)

full_r2 <- r2(m1)$R2_marginal

drop_var <- function(var) {
  formula_new <- update(formula(m1), paste(". ~ . -", var))
  model_new <- lmer(formula_new, data = dat)
  full_r2 - r2(model_new)$R2_marginal
  }

vars <- c("age_yrs", "gender", "cc", "CaseControl",
          "new_Bcell", "new_CD4T", "new_CD8T", "new_Mono", "new_NK")

sapply(vars, drop_var) %>% 
  enframe(name = "variable", value = "R2 drop") %>% 
  arrange(desc(`R2 drop`))
```
Negative values means R2 went up when variable was removed.

## Clustering Validation

Filter spline model w/CpGs to test
pred_age = the expected M-value (log2 ratio) at each age
  * change in log-ratio per unit age for first differences
  * how that log-ratio change itself is changing for second differences
  
  
### Defining "Gold Standard" CpGs 

```{r}
time_range <- max(pheno_complete$age_yrs) - min(pheno_complete$age_yrs)

cpgs_to_test2 <- lmeR_MAD_results %>%
  mutate(
    baseline_beta = m2beta(fixed_intercept),
    end_m = fixed_intercept + fixed_slope * time_range,
    end_beta = m2beta(end_m),
    delta_beta = end_beta - baseline_beta,
    group = case_when(
      fixed_slope > 0.01 &
        baseline_beta < 0.8 &
        delta_beta > 0.2 ~ "increasing",

      abs(fixed_slope) < 0.01 &
        abs(delta_beta) < 0.05 ~ "flat",
        #baseline_beta > 0.1 & #flat CpGs that had room to change but didnt
        #baseline_beta < 0.9 ~ "flat", #flat CpGs that had room to change but didnt
        #var.rand.slope < 0.01 ~ "flat", # essentially zero var btw subjects

      fixed_slope < -0.01 &
        baseline_beta > 0.2 &
        delta_beta < -0.2 ~ "decreasing",
      TRUE ~ NA_character_)) %>%
  filter(!is.na(group) & is_singular!=TRUE) %>%
  group_by(group) %>%
  #slice_head(n = 1000) %>%
  ungroup() %>%
  select(-c(starts_with("intercept_"), starts_with("slope_")))
  
```


```{r}
lmeR_MAD_results %>%
  mutate(
    baseline_beta = m2beta(fixed_intercept),
    end_m = fixed_intercept + fixed_slope * time_range,
    end_beta = m2beta(end_m),
    delta_beta = end_beta - baseline_beta) %>% 
  select(delta_beta) %>% 
  summary()

cpgs_to_test2 %>% 
  group_by(group) %>% 
  summarise(n = n())
```

Get the empirical betas for a specific CpG at specific age (time)
```{r}
# example CpG
probe <- "cg20789941"
i <- match(probe, probe_ids)

if (is.na(i)) stop("Probe not found: ", probe)

pheno <- pheno_complete[
  abs(pheno_complete$age_yrs - target_age) <= window,
  c("rgName", "maskid", "age_yrs")]

# pull M-values for one CpG
cpg <- as.numeric(matrix.clean[i, ])
names(cpg) <- sample_ids

# attach M-values to phenotype by sample ID
pheno$M_value <- cpg[as.character(pheno$rgName)]

# convert observed M-values to Beta-values
pheno$Beta_value <- m2beta(pheno$M_value)

age <- 1.0
window <- 0.25
# over observations - not subjects
pheno %>%
  filter(abs(age_yrs - age) <= window) %>%
  summarise(
    n_obs = n(),
    mean_age = mean(age_yrs, na.rm = TRUE),
    mean_beta = mean(Beta_value, na.rm = TRUE),
    sd_beta = sd(Beta_value, na.rm = TRUE))
```


```{r}
# model diagnostic differences between LMM w/ random int/slope vs
# model with nonlinear slope and linear random int/slope

aic_comp <- lmeR_MAD_results %>% # linear age
  select(CpG, aic_mad = aic, bic_mad = bic) %>%
  inner_join(
    lmeR_spline_int_res %>% # non-linear age
      select(CpG, aic_spline = aic, bic_spline = bic),
    by = "CpG") %>%
  mutate(
    aic_diff = aic_mad - aic_spline,
    bic_diff = bic_mad - bic_spline) %>% 
  column_to_rownames(var = "CpG")
  
```

### Clustering Features

```{r}
test_res <- lmeR_spline_int_res[lmeR_spline_int_res$CpG %in% cpgs_to_test2$CpG,]

# extract trajectory matrix
pred_cols <- grep("^pred_age_", names(test_res), value = TRUE)
# extract other features of interest
ot_cols <- c("fixed_intercept", "se_intercept", "t_intercept",
             "ns1_se", "ns2_se", "ns3_se", "ns1_t", "ns2_t", "ns3_t",
             "var.rand.int", "var.rand.slope", "var.resid")

features <- c(pred_cols, ot_cols)

X.traj <- as.matrix(test_res[, ..pred_cols])
rownames(X.traj) <- test_res$CpG
```


Extract and sort projected curves
```{r}
# extract numeric ages
ages <- as.numeric(sub("pred_age_", "", pred_cols))

# order columns by age
ord <- order(ages)
pred_cols <- pred_cols[ord]
ages <- ages[ord]
```


```{r}
# combine other features into clustering algo
common_cpgs <- Reduce(intersect, list(rownames(X.traj),rownames(aic_comp)))

traj_mat_test <- cbind(
  X.traj[common_cpgs, , drop = F],
  # cpgs_to_test2[cpgs_to_test2$CpG %in% common_cpgs, 
  #               c("baseline_beta", "delta_beta")],
  aic_comp[common_cpgs, c("aic_diff", "bic_diff"), drop = F])
```


```{r}
# !colnames(traj_mat_test) %in% c("pred_age_3.724", "pred_age_6.190")
# scale
traj_scaled <- scale_rows(traj_mat_test[, 1:12])
X_scaled <- scale(X.traj)

# need column wide scaling for non-trajectory features
#ot_scaled <- scale()
```

### PCA

```{r}
# PCA
pca_test <- prcomp(traj_shape_test, center = FALSE, scale. = FALSE)
summary(pca_test)
pc_use_test <- pca_test$x[, 1:3] # covers 100% variability
```

```{R}
# PCA loadings = corr btw variables and PC = how much each variable contributes to
pca_test$rotation[,1] %>% .[order(abs(.), decreasing = T)] %>% 
  enframe(name = "variable", value = "loading")

pca_test$rotation[,2] %>% .[order(abs(.), decreasing = T)] %>% 
  enframe(name = "variable", value = "loading")

pca_test$rotation[,3] %>% .[order(abs(.), decreasing = T)] %>% 
  enframe(name = "variable", value = "loading")

```


### HC
```{r}
# hierarchical clustering
dist_mat <- dist(X_scaled)
hc <- hclust(dist_mat, method = "ward.D2")
pred <- cutree(hc, k = 3)
cpgs_to_test2$pred <- pred

# correlation distance
cor_dist <- as.dist(1 - cor(t(X_scaled), use = "pairwise.complete.obs"))
hc <- hclust(cor_dist, method = "average")
pred <- cutree(hc, k = 3)
cpgs_to_test2$pred <- pred
```


### K-Means

```{r}
set.seed(2522)
km <- kmeans(traj_scaled, centers = 3, nstart = 50)
```

CpG1 = (3.6, 3.2, 2.8)
CpG2 = (2.3, 2.3, 2.2)
kmeans clusters based on distance between these row vectors across pred_**

```{r}
pred <- km$cluster[match(cpgs_to_test2$CpG, 
                         rownames(X_scaled))] #rownames obj needs to change
cpgs_to_test2$pred <- pred
```


```{r}
adjustedRandIndex(cpgs_to_test2$group, pred)
cbind(cpgs_to_test2$group, pred)
```

```{r}
# summarize K-means classification
km_class <- data.frame(
  CpG = cpgs_to_test2$CpG,
  og_group = cpgs_to_test2$group,
  pred_cluster = pred,
  pred_group = case_when(
    pred == 2 ~ "increasing", # **PREDICTED LABELS CHANGE EACH KM RUN**
    pred == 3 ~ "decreasing",
    pred == 1 ~ "flat"
  )) %>%
  mutate(is_correct = case_when(
    og_group == "decreasing" & pred_cluster == 3 ~ TRUE,
    og_group == "increasing" & pred_cluster == 2 ~ TRUE,
    og_group == "flat" & pred_cluster == 1 ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  remove_rownames()

# misclassified classes
cpg_wrg <- km_class %>% filter(!is_correct) %>% pull(CpG)

```


### View CpG trajectories

```{r}
# randomly look at cpgs
getSpagPlot_big(
  CpG = "cg26928153",
  mvals_bm = matrix.clean,
  value_type = "beta",
  y_limits = c(0.85, 1),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  #group_col = "gender",
  #facet_by_group = T,
  legend = "none")

```


```{r}
# using only latent curves
suppressWarnings( # Warnings relate to missing values which I am aware of
for (i in cpg_diff$CpG[1:50]){
  p <- getSpagPlot_big(
  CpG = i,
  mvals_bm = matrix.clean,
  pheno = pheno_complete,
  anno = anno,
  value_type = "beta", #beta or m
  y_limits = c(0,1),
  age_unit = "years",
  probe_ids = probe_ids,
  sample_ids = sample_ids)
  print(p)
})

lmeR_spline_int_res[lmeR_spline_int_res$nObs==730,]
```


```{r}
# using only latent curves
suppressWarnings( # Warnings relate to missing values which I am aware of
for (i in cpg_wrg){
  p <- getSpagPlot_big(
  CpG = i,
  mvals_bm = matrix.clean,
  pheno = pheno_complete,
  anno = anno,
  value_type = "beta", #beta or m
  y_limits = c(0,1),
  probe_ids = probe_ids,
  sample_ids = sample_ids)
  print(p)
})

# library(patchwork)
# plots <- lapply(cpg_wrg, function(i) {
#   getSpagPlot_big(
#     CpG = i,
#     mvals_bm = matrix.clean,
#     pheno = pheno_complete,
#     anno = anno,
#     value_type = "m", #beta or m
#     #y_limits = c(0,1),
#     probe_ids = probe_ids,
#     sample_ids = sample_ids)
#   })
# 
# wrap_plots(plots, ncol = 2)

```

```{r}
# data table method
# cpgs_to_test2[CpG %in% cpg_wrg,
#               .(CpG, nObs, fixed_intercept, fixed_slope, baseline_beta, 
#                    delta_beta, group,is_singular)]

# Turn off scientific notation
options(scipen = 999)
cpgs_to_test2 %>%
  filter(CpG %in% cpg_wrg) %>%
  left_join(km_class %>% select(CpG, pred_group), by = "CpG") %>%
  select(CpG, nObs, fixed_slope, baseline_beta, 
         delta_beta, group, pred_group, var.rand.slope,is_singular) %>% 
    mutate(across(c(fixed_slope, delta_beta, var.rand.slope), 
                ~ round(., 6))) 
# Reset to default when done
options(scipen = 0)

```

### Review mislabeled CpGs

```{r}
# Data.Table method
# cpgs_to_test2[CpG %in% c("cg19159092", "cg01729262"),
#               .(CpG, nObs, fixed_intercept, fixed_slope, baseline_beta, 
#                    delta_beta, group,is_singular)]

# Turn off scientific notation
options(scipen = 999)
cpgs_to_test2 %>%
  filter(CpG %in% c("cg19159092", "cg01729262")) %>%
  #left_join(km_class %>% select(CpG, pred_group), by = "CpG") %>%
  select(CpG, nObs, fixed_slope, baseline_beta, 
         delta_beta, group, var.rand.slope,is_singular) %>% 
    mutate(across(c(fixed_slope, delta_beta, var.rand.slope), 
                ~ round(., 6))) 
# Reset to default when done
options(scipen = 0)
```

### Plot Fitted Trajectories from Spline Model

```{r}
# two trajectory sites: cg03013609 ; cg20239921
# well behaved sites: cg23776628 ; 
cpg_i <- "cg12025418" 

pred_traj_plot <- make_spline_pred_df(
  CpG = cpg_i,
  fitted_model = lmeR_spline_int_full, #lmeR_spline_int_slope_res
  value_type = "beta")

getSpagPlot_big(
  CpG = cpg_i,
  mvals_bm = matrix.clean,
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  pred_df = pred_traj_plot,
  #y_limits = c(0,1),
  value_type = "m",
  age_unit = "years") # or days

```


### Derivatives from Predicted Trajectories
 
$$
D1_k = \frac{\hat{M}(a_{k+1}) - \hat{M}(a_k)} {a_{k+1} - a_k}
$$

```{r}
# change between time points
# age grid is uneven so we need to divide by age gap
# X.traj = predicted trajectory matrix
# rows = CpGs, columns = predicted ages

# extract trajectory column names
pred_cols <- grep("^pred_age_", names(spline_int_results), value = TRUE)
# extract numeric ages
ages <- as.numeric(sub("pred_age_", "", pred_cols))

# order columns by age
ord <- order(ages)
pred_cols <- pred_cols[ord]
ages <- ages[ord]

# predicted M-value matrix: rows = CpGs, columns = ages
Mhat <- as.matrix(spline_int_results[, ..pred_cols])
rownames(Mhat) <- spline_int_results$CpG

# identify CpGs with complete predicted trajectories
# CpGs with contrast error identified earlier - e.g. class misbalance
good_cpgs <- rowSums(is.na(Mhat)) == 0
Mhat <- Mhat[good_cpgs, , drop = FALSE]

# convert predicted M-values to Beta-values
Betahat <- m2beta(Mhat)
rm(Mhat) # don't need Mhat matrix

# numerical derivative dBeta/dage between adjacent ages
dBeta_dage <- t(apply(Betahat, 1, function(x) {
  diff(x) / diff(ages)
  }))

# name derivative columns by age interval
colnames(dBeta_dage) <- paste0(
  "dBeta_dage_",
  round(ages[-length(ages)], 3),
  "_to_",
  round(ages[-1], 3))

```

#### Identifying Transient Trajectories (Change point analysis)


Filtering translation:
predicted methylation changes by less than 1 percentage point across the full age range.
no local interval changes faster than 0.5 percentage points per year.

> dBeta_dage_0.203_to_0.501 = estimated change in predicted Beta-value per 1 year of age between age 0.994 and 1.500
>
> Predicted methylation Beta-value increases by about 0.00179 per year over that
age interval for cg26928153.
>
> With splines, there is no single age slope. Instead, the slope changes depending
on where you are along the age trajectory. This derivative table gives the local 
rate of change between adjacent ages.
>
> We are taking differences between adjacent predicted methylation values along
the fitted trajectory and dividing by the age interval to estimate the local 
rate of methylation change. This is an approximation to dBeta/dage - 
the local rate of methylation change along the spline trajectory.

Tells us:
* how M-scale slopes translate to Beta-scale slopes
* why the Beta-scale slope depends on the current Beta value

#### K-means on "non-flat" CpGs

Parallelization

```{r}
set.seed(2522)
Ks <- 2:7
# scale trajectories
traj_scaled <- scale_rows(Betahat.red)

# sample CpGs for silhouette evaluation
sil_idx <- sample(seq_len(nrow(traj_scaled)), 10000)
traj_sil <- traj_scaled[sil_idx, , drop = FALSE]

plan(multisession, workers = 12)

t0 <- Sys.time() # Start time

sil_vals <- future_sapply(Ks, function(k) {
  set.seed(2522 + k)
  fit <- kmeans(
    traj_scaled, 
    centers = k, 
    nstart = 25, 
    iter.max = 1000,
    algorithm = "Hartigan-Wong")
  # replicate a few times
  sil_rep_vals <- replicate(5, { # do 5 times to start
    idx <- sample(seq_len(nrow(traj_scaled)), 10000) # test 10k CpGs
    sil <- silhouette(fit$cluster[idx], dist(traj_scaled[idx, , drop = F]))
    mean(sil[, 3])
    })
  mean(sil_rep_vals)}, 
  future.seed = T)

plan(sequential) # release the cpu workers

t1 <- Sys.time() # End time
message("Silhouette runtime: ", 
        round(as.numeric(difftime(t1, t0, units = "mins")), 2), 
        " minutes")

(best_k <- Ks[which.max(sil_vals)])

plot(
  Ks, sil_vals,
  type = "b",
  pch = 19,
  xlab = "Number of Clusters (k)",
  ylab = "Average Silhouette Width",
  main = "Optimal k by Sampled Silhouette Score")

abline(v = best_k, col = "red", lty = 2)

```

```{r}
kmeans_grid <- function(x, k_values = 3:12, nstart = 20, iter.max = 100) {
  out <- future_lapply(k_values, function(k) {
    message("Running kmeans for k = ", k)
    km <- kmeans(x, centers = k, nstart = nstart, iter.max = iter.max,
                 algorithm = "Hartigan-Wong")
    list(
      k = k,
      km = km,
      tot_withinss = km$tot.withinss,
      betweenss = km$betweenss,
      totss = km$totss,
      ratio = km$betweenss / km$totss)
  }, future.seed = TRUE)
  
  names(out) <- paste0("k", k_values)
  out
}

# sum of squares (SS) method
km_grid <- kmeans_grid(
  x = traj_scaled,
  k_values = 2:10,
  nstart = 25,
  iter.max = 1000)

km_summary <- rbindlist(
  lapply(km_grid, function(x) {
    data.table(
      k = x$k,
      tot_withinss = x$tot_withinss,
      betweenss = x$betweenss,
      totss = x$totss,
      ratio = x$ratio)
  }))

print(km_summary)
```



```{r}
# run Kmeans
set.seed(2522)
km <- kmeans(traj_scaled, centers = best_k, nstart = 50)

pred <- km$cluster
pred_vector <- pred[match(rownames(Betahat.red), names(pred))]
table(km$cluster)

# cluster_df <- data.frame(
#   CpG = rownames(Betahat.red),
#   pred = pred_vec)
```

```{r}
pca <- prcomp(traj_scaled)
summary(pca)

set.seed(2522)
plot_idx <- sample(seq_len(nrow(traj_scaled)),10000)
plot(
  pca$x[plot_idx, 1],
  pca$x[plot_idx, 2],
  col = km$cluster[plot_idx],
  pch = 16,
  cex = 0.5)

```

```{r}
cluster_means <- rowsum(traj_scaled, group = km$cluster) / 
  as.vector(table(km$cluster))

matplot(
  ages,
  t(cluster_means),
  type = "l",
  lty = 1,
  lwd = 3,
  xlab = "Age",
  ylab = "Mean Beta",
  main = "Cluster-average methylation trajectories")

legend(
  "topright",
  legend = paste("Cluster", rownames(cluster_means)),
  col = 1:nrow(cluster_means),
  lty = 1,
  lwd = 3)
```


```{r}
pca$rotation[,1] %>% .[order(abs(.), decreasing = T)] %>% 
  enframe(name = "variable", value = "loading")

pca$rotation[,2] %>% .[order(abs(.), decreasing = T)] %>% 
  enframe(name = "variable", value = "loading")

```


```{r}
plot(
  ages,
  pca$rotation[,1],
  type = "b",
  pch = 19,
  xlab = "Age",
  ylab = "PC1 loading",
  main = "PC1 trajectory pattern")

plot(
  ages,
  pca$rotation[,2],
  type = "b",
  pch = 19,
  xlab = "Age",
  ylab = "PC2 loading",
  main = "PC2 trajectory pattern")
```


ARI measues how well pairs of points are grouped
* compares predicted clusters with true labels
	*	1.0 → perfect agreement (clusters match exactly)
	*	0.0 → random clustering (no better than chance)
	*	< 0 → worse than random
	
	
* "Cluster A shows increasing methylated/unmethylated signal with age”
* “Cluster B shows decreasing relative methylation levels”
* “Cluster C shows stable trajectories”

## GMM

Identify sub-trajectories 
```{r}
# remove flat CpGs
# spline_intercepts <- 
#   spline_int_BLUPs[!spline_int_BLUPs$CpG %in% rownames(flat_df),]

# extract intercept columns
int_cols <- grep("^intercept_", names(spline_int_BLUPs), value = TRUE)
# fit GMM
gmm.model <- Mclust(spline_intercepts[, ..int_cols][1:50], G = 1:3)
summary(gmm.model, parameters = TRUE)
gmm.model$modelName

gmm.model$df
str(gmm.model)
# how many borderline points
sum(apply(gmm.model$z, 1, max) <0.9)

# most uncertain row
which.min(apply(gmm.model$z, 1, max))
spline_intercepts[441,]
head(round(gmm.model$z,3))

# uncertain rows
unc_idx <- which(apply(gmm.model$z, 1, max) < 0.95)
head(round(gmm.model$z[unc_idx,],3),6)
gmm.model$classification
```


```{r}
# run GMM
options(future.globals.maxSize = 5 * 1024^3)
plan(multicore, workers = 12)

gmm_results <- future_lapply(
  seq_len(nrow(spline_intercepts)),
  function(i) {
    
    x <- as.numeric(spline_intercepts[i, ..int_cols])
    x <- x[is.finite(x)]
    
    # --- insufficient data ---
    if (length(x) < 20) {
      return(data.table(
        CpG = spline_intercepts$CpG[i],
        best_G = NA_integer_,
        bic = NA_real_,
        mean_uncertainty = NA_real_,
        max_uncertainty = NA_real_,
        min_prop = NA_real_,
        mean_sep_min = NA_real_,
        sd_min = NA_real_,
        sd_max = NA_real_,
        call_type = "insufficient_data"))
    }
    
    # --- near-zero variance -> trivial group 1 ---
    if (sd(x) < 1e-8) {
      return(data.table(
        CpG = spline_intercepts$CpG[i],
        best_G = 1L,
        bic = NA_real_,
        mean_uncertainty = 0,
        max_uncertainty = 0,
        min_prop = 1,
        mean_sep_min = NA_real_,
        sd_min = sd(x),
        sd_max = sd(x),
        call_type = "zero_variance"))
    }
    
    mod <- tryCatch(
      Mclust(x, G = 1:3, verbose = FALSE),
      error = function(e) NULL)
    
    # --- Mclust failed ---
    if (is.null(mod)) {
      return(data.table(
        CpG = spline_intercepts$CpG[i],
        best_G = NA_integer_,
        bic = NA_real_,
        mean_uncertainty = NA_real_,
        max_uncertainty = NA_real_,
        min_prop = NA_real_,
        mean_sep_min = NA_real_,
        sd_min = NA_real_,
        sd_max = NA_real_,
        call_type = "mclust_error"))
    }
    
    means <- sort(as.numeric(mod$parameters$mean))
    props <- as.numeric(mod$parameters$pro)
    sds <- tryCatch(
      sqrt(as.numeric(mod$parameters$variance$sigmasq)),
      error = function(e) NA_real_)
    
    # --- normal fitted case ---
    data.table(
      CpG = spline_intercepts$CpG[i],
      best_G = mod$G,
      bic = max(mod$BIC, na.rm = TRUE),
      mean_uncertainty = mean(mod$uncertainty, na.rm = TRUE),
      max_uncertainty = max(mod$uncertainty, na.rm = TRUE),
      min_prop = min(props, na.rm = TRUE),
      # for bi/tri-modal groups only (G>1)
      # tells use how spread-out the subjected are within a Cpg's fitted component
      mean_sep_min = if (mod$G > 1) min(diff(means)) else NA_real_,
      sd_min = min(sds, na.rm = TRUE),
      sd_max = max(sds, na.rm = TRUE),
      call_type = "fitted")
  },
  future.seed = TRUE)

gmm_results <- data.table::rbindlist(gmm_results, fill = TRUE)
```



```{r}
# qsave(gmm_results, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/gmm_results.qs", nthreads = 12)

# GMM results w/chrX removed and controls only
gmm_results <- qread(
  "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/gmm_filtered_results.qs", 
  nthreads = 12)

```

```{r}
gmm_results |> group_by(best_G) |> summarise(n=n())
unimodal_cpgs <- gmm_results[gmm_results$best_G==1, "CpG"]
bimodal_cpgs <- gmm_results[gmm_results$best_G==2, "CpG"]
trimodal_cpgs <- gmm_results[gmm_results$best_G==3, "CpG"]
# unimodal_cpgs <- qsave(unimodal_cpgs, 
#       "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/unimodal_cpgs.qs", 
#       preset = "fast", nthreads = 12)
```

Filtering uncertainty out of GMM results
```{r}
bimodal_cpgs <- gmm_results[
  best_G == 2 & mean_uncertainty < 0.1 & min_prop > 0.1 & mean_sep_min > 0.5]

trimodal_cpgs <- gmm_results[
  best_G == 3 & mean_uncertainty < 0.1 & min_prop > 0.1 & mean_sep_min > 0.5]

cpg_unc <- bimodal_cpgs[bimodal_cpgs$max_uncertainty < 0.4, "CpG"]
gmm_grp1_cpgs <- gmm_results[gmm_results$best_G==1, "CpG"]
```


```{r}
gmm_results |> filter(call_type=="zero_variance")
```

```{r}
table(gmm_results_filtered$call_type)
```

Calculate SDs for intercept BLUPs to determine CpG site that have low variability
between subjects
```{r}
blup_sd <- apply(
  spline_int_BLUPs[, ..int_cols], 1, function(x) 
    {
    x <- as.numeric(x)
    x <- x[is.finite(x)]
    if (length(x) < 2) return(NA_real_)
    sd(x)
  })

table(blup_sd < 1e-6, useNA = "always")
summary(blup_sd)
```

"CpGs with negligible inter-individual variability in random-intercept BLUPs, for which Gaussian mixture modeling was not applicable"

```{r}
summary(blup_sd[blup_sd < 1e-6])
quantile(blup_sd[blup_sd < 1e-6], probs = c(0, .25, .5, .75, 1))
```

```{r}
spline_summary_BLUPs[spline_summary_BLUPs$CpG=="cg04172501",]
```


```{r}
# tri-CpGs c(cg10197305, cg23692511, cg14531093)
getSpagPlot_big(
  CpG = "cg04172501",
  mvals_bm = matrix.clean,
  value_type = "beta",
  #y_limits = c(0.05, .2),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  #group_col = "CaseControl",
  #facet_by_group = T,
  legend = "bottom")

cpg_nas <- gmm_results[gmm_results$call_type=="zero_variance",]
grp1_gmm <- gmm_results[gmm_results$best_G==1, ]
for (i in sample(grp1_gmm$CpG, 50)) {
  p <- getSpagPlot_big(
    CpG = i,
    mvals_bm = matrix.clean,
    value_type = "beta",
    y_limits = c(0, 1),
    age_unit = "years",
    pheno = pheno_complete,
    anno = anno,
    probe_ids = probe_ids,
    sample_ids = sample_ids,
    #group_col = "HLA_Category",
    #facet_by_group = T,
    legend = "bottom")
  print(p)
  }
```

"The NAs identified by GMM (if (sd(x) < 1e-8)) appear to have really tight
starting methylation values.

### GMM Visualized

```{r}
tri_ord <- trimodal_cpgs %>% arrange(mean_uncertainty)
for (i in tri_ord$CpG[1:50]) {
  p <- getSpagPlot_big(
    CpG = i,
    mvals_bm = matrix.clean,
    value_type = "beta",
    y_limits = c(0, 1),
    age_unit = "years",
    pheno = pheno_complete,
    anno = anno,
    probe_ids = probe_ids,
    sample_ids = sample_ids,
    #group_col = "HLA_Category",
    #facet_by_group = T,
    legend = "bottom")
  print(p)
  }

```

```{r}
# tri-CpGs c(cg10197305, cg23692511, cg14531093)
getSpagPlot_big(
  CpG = "cg04172501",
  mvals_bm = matrix.clean,
  value_type = "beta",
  #y_limits = c(0.05, .2),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  #group_col = "CaseControl",
  #facet_by_group = T,
  legend = "bottom")
```

View histogram of GMM. ideally there should be multiple peaks (2 or 3)

```{r}
plot_gmm_hist(
  cpg = "cg01014490", 
  spline_intercepts=spline_int_BLUPs, # intercepts from spline model
  int_cols, # intercept column names
  legend_pos = "topleft",
  show_empirical = FALSE)
```


```{r}
for (i in tri_ord$CpG[1:50]) {
  p <- plot_gmm_hist(
  cpg = i, 
  lmeR_spline_int_BLUP[CpG==i,], # intercepts from spline model
  int_cols, # intercept column names
  legend_pos = "topleft",
  show_empirical = FALSE
  )
  print(p)
}
```


* Solid colored lines = GMM model components
* Single dashed black line = empirical density (reference only)
* Colored fill = under the model, not the empirical curve


## Transient/Change point Analysis

```{r}
# change between time points
# age grid is uneven so we need to divide by age gap
# X.traj = predicted trajectory matrix
# rows = CpGs, columns = predicted ages

# extract trajectory column names
pred_cols <- grep("^pred_age_", names(spline_summary_BLUPs), value = TRUE)
# extract numeric ages
ages <- as.numeric(sub("pred_age_", "", pred_cols))

# order columns by age
ord <- order(ages)
pred_cols <- pred_cols[ord]
ages <- ages[ord]

# predicted M-value matrix: rows = CpGs, columns = ages
Mhat_traj <- as.matrix(spline_summary_BLUPs[, ..pred_cols])
rownames(Mhat_traj) <- spline_summary_BLUPs$CpG

# identify CpGs with complete predicted trajectories
# CpGs with contrast error identified earlier - e.g. class misbalance
good_cpgs <- rowSums(is.na(Mhat_traj)) == 0
Mhat_traj <- Mhat_traj[good_cpgs, , drop = FALSE]

# convert predicted M-values to Beta-values
Betahat_traj <- m2beta(Mhat_traj)
#rm(Mhat_traj) # don't need Mhat matrix

# numerical derivative dBeta/dage between adjacent ages
# Beta-scale derivatives
dBeta_dage <- t(apply(Betahat_traj, 1, function(x) {diff(x) / diff(ages)} ))

# M-value-scale derivatives
dM_dage <- t(apply(Mhat_traj, 1, function(x) {diff(x) / diff(ages)} ))

# name derivative columns by age interval
colnames(dBeta_dage) <- paste0(
  "dBeta_dage_",
  round(ages[-length(ages)], 3),
  "_to_",
  round(ages[-1], 3))
```

```{r}
m_start <- Mhat_traj[, 1]; m_end <- Mhat_traj[, ncol(Mhat_traj)]

m_endpoint_change <- m_end - m_start

m_range <- rowMaxs(Mhat_traj) - rowMins(Mhat_traj)

mean_abs_m_slope <- rowMeans(abs(dM_dage))
max_abs_m_slope <- rowMaxs(abs(dM_dage))
```

```{r}
plot(Mhat_traj[, 1], Mhat_traj[, ncol(Mhat_traj)],
  xlab = "Predicted M-value at earliest age",
  ylab = "Predicted M-value at latest age",
  pch = 16, cex = 0.2)
abline(0, 1, col = "red")
```

```{r}
hist(m_range, breaks = 100,
     xlim = c(0, 1),
     main = "Predicted M-value trajectory range",
     xlab = "Maximum minus minimum predicted M-value")

hist(mean_abs_m_slope, breaks = 100,
  main = "Mean absolute M-value slope",
  xlab = "M-value units per year")

summary(mean_abs_m_slope)

# absolute change so this will capture any positive or negative change together****
hist(mean_abs_m_slope,breaks = 100,
  xlim = c(0, 0.2),
  main = "Mean absolute M-value slope",
  xlab = "M-value units per year")
```



```{r}
interval_mid   <- (ages[-length(ages)] + ages[-1]) / 2
interval_width <- diff(ages)

traj_range_m <- apply(Mhat_traj, 1, function(x) max(x) - min(x))
total_abs_change_m <- rowSums(abs(dM_dage) *
                               matrix(interval_width, nrow = nrow(dM_dage),
                                      ncol = length(interval_width), byrow = TRUE))
monotonicity_gap <- total_abs_change_m - traj_range_m
```

```{r}
## does the gap correlate with overall magnitude (traj_range), or is it independent?
plot(traj_range_m, monotonicity_gap, pch = ".", 
     xlab = "Range (M-units)", ylab = "TV - Range (monotonicity gap)")
```



Segmented Analysis to determine a breakpoint for "early" vs "late" split
```{r}
# breakpoint analysis
fit_lin_fine  <- lm(median_abs ~ interval_mid, data = slope_summary_fine)
fit_seg_fine  <- segmented(fit_lin_fine, seg.Z = ~interval_mid,
                            psi = list(interval_mid = 2))

summary(fit_seg_fine)
fit_seg_fine$psi
```

Confirm it's not just locally trapped near the seed
```{r}
starts <- c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4)
results <- lapply(starts, function(s) {
  tryCatch({
    fit <- segmented(fit_lin_fine, seg.Z = ~interval_mid, psi = list(interval_mid = s))
    c(start = s, psi = fit$psi[1, "Est."], se = fit$psi[1, "St.Err"])
  }, error = function(e) c(start = s, psi = NA, se = NA))
})
do.call(rbind, results)
```

Boostrap to get accurate SE. The grid points are not independent observations. They're
deterministic evaluations of the same underlying spline curve.

```{r}
library(parallel)

n_boot <- 500
n_cores <- 12  

run_one_boot <- function(b, dM_dage_fine, fine_interval_mid) {
  n_cpg <- nrow(dM_dage_fine)
  idx <- sample(seq_len(n_cpg), n_cpg, replace = TRUE)
  boot_median <- apply(abs(dM_dage_fine[idx, , drop = FALSE]), 2, median, na.rm = TRUE)

  df_b <- data.frame(interval_mid = fine_interval_mid, median_abs = boot_median)
  fit_b <- tryCatch(
    segmented(lm(median_abs ~ interval_mid, data = df_b),
              seg.Z = ~interval_mid, psi = list(interval_mid = 2)),
    error = function(e) NULL
  )
  if (is.null(fit_b)) return(NA_real_)
  fit_b$psi[1, "Est."]
}

set.seed(2522) 
psi_boot <- unlist(mclapply(seq_len(n_boot), run_one_boot,
                             dM_dage_fine = dM_dage_fine,
                             fine_interval_mid = fine_interval_mid,
                             mc.cores = n_cores))

mean(is.na(psi_boot))
quantile(psi_boot, c(0.025, 0.5, 0.975), na.rm = T)
summary(psi_boot)
# qsave(psi_boot, 
#       "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/psi_boot.qs", 
#       nthreads=12)
```

What segmented() is actually telling you:

"It answers one narrow question: at what age does the population-average |dM/dage| curve stop declining and level off? That's a statement about the aggregate trend across all CpGs pooled together. The 3.58y estimate says "on average, across your whole genome/panel, methylation dynamics have mostly settled down by ~3.6 years."

```{r}
davies.test(fit_lin, seg.Z = ~interval_mid)
```

```{r}
## Range of predicted M-values across the whole trajectory, per CpG
traj_range <- apply(traj_matrix, 1, function(x) max(x) - min(x))

## Or: total variation of the derivative-integrated curve (equivalent-ish, more robust to monotonic vs non-monotonic shapes)
total_abs_change <- rowSums(
  abs(dM_dage_fine) * matrix(fine_interval_width,
                             nrow = nrow(dM_dage_fine),
                             ncol = length(fine_interval_width), byrow = T))

hist(traj_range, breaks = 100)
summary(total_abs_change)
hist(total_abs_change, breaks=100)

hist(log10(total_abs_change), breaks = 100,
     main = "log10(total absolute M-value change)")
```

"With a smooth unimodal distribution like this, wherever you draw a percentile line, you're bisecting a continuum at a point with no distributional justification — the CpGs just below your cutoff and just above it are barely distinguishable from each other. That's not "slightly less rigorous," it's actually undefined what you're separating. Any threshold here needs to be justified by something external to this distribution — either a statistical criterion or a biological benchmark — not by the shape of the histogram itself.


```{r}
# total absolute change in beta
total_abs_beta_change <- rowSums(abs(dBeta_dage) * diff(ages))

# OR explicitly:
total_abs_beta_change <- rowSums(
  abs(dBeta_dage) * matrix(diff(ages), 
                             nrow = nrow(dBeta_dage), 
                             ncol = ncol(dBeta_dage), 
                             byrow = TRUE))

# 3. Histograms
par(mfrow = c(2, 2))  # 2x2 layout

hist(Betahat_traj, breaks = 100, 
     main = "Range of Beta across trajectory",
     xlab = "Beta range", col = "lightblue")

hist(total_abs_beta_change, breaks = 100,
     main = "Total absolute Beta change",
     xlab = "Total variation", col = "lightgreen")

hist(log10(total_abs_beta_change + 1e-10), breaks = 100,  # Added small constant to avoid -Inf
     main = "log10(total absolute Beta change)",
     xlab = "log10(total variation)", col = "lightyellow")

# Summary stats
cat("Range Beta:\n")
summary(Betahat_traj)
cat("\nTotal absolute Beta change:\n")
summary(total_abs_beta_change)

par(mfrow = c(1, 1))  # Reset layout
```

Determine % of CpG sites with
```{r}
thresholds <- c(0.02, 0.05, 0.07, 0.10, 0.15)
frac_above <- sapply(thresholds, function(x) mean(total_abs_beta_change > x))
names(frac_above) <- paste0(">", thresholds*100, "%")
round(frac_above,3)
```

"CpGs were retained for functional PCA if their fitted β-value trajectory spanned ≥3% across the study period. This threshold yielded a dynamic fraction (~X%) comparable to the ~14% of CpGs identified as significantly age-associated in an overlapping early-childhood window by Pérez et al. (2019)." That's a much stronger justification than an internally-derived percentile, since it's anchored to an external, independent estimate of what fraction of the methylome is genuinely dynamic in this age range — and now you have the actual number to back it up rather than just an analogy.

You compute both traj_range_beta (max − min) and total_abs_beta_change. For a strictly monotonic curve, these are mathematically equal (total variation = endpoint-to-endpoint range). For a non-monotonic curve (rise-then-fall, i.e., a transient bump), total variation will exceed the range, because it separately sums the "up" leg and the "down" leg rather than netting them out.


## Flat Filtering

Genomic inflation???***
```{r}
spline_summary_BLUPs |> 
  filter(age_omnibus_p < 0.05) |> 
  summarise(n = n())
```


## dpGMM

```{r}
cpg <- spline_intercepts$CpG[1]

x <- as.numeric(spline_intercepts[CpG == cpg, ..int_cols])
x <- x[is.finite(x)]

opts <- GMM_1D_opts()
str(opts)
```

Pick 25-50 w/visually confirmed sub-trajectories and compare McClust w/ dpGMM

```{r}
test_cpgs <- tri_ord$CpG[1:50]

opts <- dpGMM::GMM_1D_opts
opts$KS <- 4
opts$IC <- "BIC"
opts$plot <- FALSE
opts$quick_stop <- FALSE

res_list <- vector("list", length(test_cpgs))

for (i in seq_along(test_cpgs)) {

  cpg <- test_cpgs[i]

  x <- lmeR_spline_int_BLUP %>%
    filter(CpG == cpg) %>%
    select(all_of(int_cols)) %>%
    unlist(use.names = FALSE) %>%
    as.numeric()

  x <- x[is.finite(x)]

  if (length(x) < 20 || sd(x) < 1e-6) next

  mod_mclust <- mclust::Mclust(x, G = 1:5, verbose = FALSE)
  mod_dpgmm  <- dpGMM::runGMM(x, opts = opts)
  
  res_list[[i]] <- data.frame(
   CpG = cpg,
   n = length(x),
   sd = sd(x),
   mclust_G = mod_mclust$G,
   mclust_BIC = max(mod_mclust$BIC, na.rm = TRUE),
   dpgmm_G = mod_dpgmm$KS,
   dpgmm_BIC = mod_dpgmm$BIC,
   dpgmm_logLik = mod_dpgmm$logLik
  )
  }
res_df <- dplyr::bind_rows(res_list)
```

```{r}
plot_gmm_hist(
  cpg = "cg16269199", 
  G_range = 1:3,
  spline_intercepts=spline_int_BLUPs, # intercepts from spline model
  int_cols, # intercept column names
  legend_pos = "topleft",
  show_empirical = FALSE
  )
```

```{r}
getSpagPlot_big(
  CpG = "cg01858500",
  mvals_bm = matrix.clean,
  value_type = "beta",
  #y_limits = c(0.05, .2),
  age_unit = "years",
  pheno = pheno_complete,
  anno = anno,
  probe_ids = probe_ids,
  sample_ids = sample_ids,
  #group_col = "CaseControl",
  #facet_by_group = T,
  legend = "bottom")
```

## RAM Usage

```{r}
data.frame(
  size_MB = round(sapply(ls(), function(x) object.size(get(x)) / 1024^2),1)) %>% 
  arrange(desc(size_MB))
```

```{r}
# determine which parts of an object use the most RAM
round(sort(
    sapply(anno, function(x) object.size(x) / 1024^2),
    decreasing = TRUE)[1:10],1)
```

lmer spline w/ RE int & slope runtime: 1548.64 minutes ≈ 26hrs
lmer spline w/ RE int only runtime: 314.61 minutes ≈ 5.2hrs


