---
title: "T1D - TEDDY Project"
author: "Ryan Summers"
date: "2026-01-06"
output:
  word_document: default
  pdf_document: default
  html_document:
    css: style.css
bibliography: t1d.bib
csl: "apa-5th-edition.csl"
nocite: '@*'
---

```{r setup, include=FALSE}
library(BiocManager)
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(kableExtra)
library(doBy)
library(MASS)
library(dplyr)
library(table1)
library(gtsummary)
library(gt)
library(stringdist)
library(ggrepel)
library(pwr)
library(ssize)
library(samr)
library(parallel)
library(BiocParallel)
library(impute)
library(limma)
library(qvalue)
library(gtools)
library(multtest)
library(GEOquery)
library(Biobase)
library(data.table)
library(SummarizedExperiment)
library(variancePartition)
library(minfi)
library(R.utils)
library(stringr)
library(methylKit)
library(shinyMethyl)
library(minfi)
library(bumphunter)
library(qs)
library(qs2)
library(lme4)
library(sesame)
library(sva)
library(EmpiricalBrownsMethod)
library(clusterProfiler)
library(patchwork)
library(ggmanh)
library(survival)
library(lme4)
library(SummarizedExperiment)
library(variancePartition)
library(lmerTest)
library(future.apply)
library(data.table)
library(splines)
# CpG Script
source("/Users/ryan_summers/GitHub/teddy-dnam-analysis/scripts/CpG_Report.R")
```

Scripts for installing multiple packages
```{r}
# quick installation script
BiocManager::install(c(
  "SummarizedExperiment",
  "BiocParallel",
  "variancePartition"))

install.packages(c(
  "", ""))
```

## Data Storage and Loading

```{r}
# TEMP for local testing
# setwd(paste0(Sys.getenv("RKJCOLLAB"), "/Maternal_Protection"))
# matrix_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata"
# sample_ids_path <- "data/teddy/TEDDYarrays/samples-for-analysis.csv"
# pheno_path <- "data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv"
# model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntAndSlopeModel.R"
# # model_path <- "/Users/slacksa/repos/explore_matprot/TEDDYcode/2-analysis/Functions/getSubSpecificInt_randomIntModel.R"
# cell_types_path <- "data/teddy/TEDDYarrays/cell proportions/cell_counts_rescaled.csv"
# cov_list <- "factor(gender),new_CD8T,new_CD4T,new_NK,new_Bcell,new_Mono"
# # cov_list <- "factor(gender)"
# cov_list <- unlist(strsplit(cov_list, ","))
```


```{r}
# # load in M-value matrix
# M_read <- load("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/mSet-norm-batch-mean-plate-row.Rdata")
# 
# # save file using serialization - quicker
# qsave(mSet.norm.mean.plate_row, "/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", preset = "fast", nthreads = 6)
```

Load data from personal external drive
```{r}
# # load in qs file
# M.matrix <- qread("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/M.matrix.qs", nthreads = 8)
# 
# qsave(M.matrix, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/M.matrix.qs", nthreads = 6)
```

850K Annotation
```{r}
# Load DNAm annotation
load(file="/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/raw/annotation.850K.Rdata")
```


## Phenotype & M-Matrix Data Management & Cleanup

```{r}
# read in phenotype
pheno <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/preprocessing-dnam/phenoData_v2.csv")
```


```{r}
# Load list of IDs to keep in analysis
sample_ids_df <- read.csv("/Volumes/Storage/Maternal_Protection/data/teddy/TEDDYarrays/samples-for-analysis.csv")
sample_ids <- sample_ids_df$rgName
```


```{r}
# Load cell type proportions
cell_types <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/cell_counts_rescaled.csv")
```


```{r}
# Filter matrix to just samples for analysis
matrix.filt <- M.matrix[, which(colnames(M.matrix) %in% sample_ids)]

# save matrix file to local
qsave(matrix.filt, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs", nthreads = 6)
```

### Load M-Matrix

```{r}
# load matrix
matrix.filt <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs", nthreads = 6)

# # save into new qs2 package
# qs_save(matrix.filt, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.filt.qs2", nthreads = 6)
```

### Load Pheno

```{r}
# Add cell types to pheno
  # TO NOTE: model should include all cell types except for neutrophils, doing
  # this essentially sets neutrophils as reference
pheno_filt_ct <- left_join(pheno_filt, cell_types,
                           by = c("Sample_Name" = "samplemaskid"))

# convert to years and center to help with different scale issue when modeling
pheno_filt_ct$age_yrs_c <-
  (pheno_filt_ct$sample_agedys / 365.25) -
  mean(pheno_filt_ct$sample_agedys / 365.25)

# scale the cell covariates
covars <- c("new_Bcell", "new_CD4T", "new_CD8T",
            "new_Mono", "new_NK")

pheno_filt_ct[covars] <- scale(pheno_filt_ct[covars])

# save to csv
write.csv(pheno_filt_ct, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_scaled.csv")
qsave(pheno_filt_ct, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_filt_ct.qs", nthreads = 6)
```


```{r}
# load in original pheno file 
pheno_filt_ct_unscaled <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_filt_ct.qs", 
                                nthreads = 6)

# pheno_filt_ct_unscaled <- read.csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno.csv",header=T)
```

```{r}
# convert age to years for scalability in model running
pheno_filt_ct_unscaled$age_yrs_c <-
  (pheno_filt_ct_unscaled$sample_agedys / 365.25) -
  mean(pheno_filt_ct_unscaled$sample_agedys / 365.25)

# convert age to years for scalability in modeul running
pheno_filt_ct_unscaled$age_yrs <- pheno_filt_ct_unscaled$sample_agedys / 365.25

# # save to csv
# write.csv(pheno_filt_ct_unscaled, file = "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno.csv")
```


```{r}
# load in pheno scaled file 
pheno_filt_ct_scaled <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/pheno_scaled.csv")
```

```{r}
# convert needed variables to factors
pheno_filt_ct_unscaled <- pheno_filt_ct_unscaled %>% 
  mutate(across(c(maskid, cc, gender, CaseControl), as.factor))


pheno_filt_ct_scaled <- pheno_filt_ct_scaled %>% 
  mutate(across(c(maskid, cc, gender, CaseControl), as.factor))
```


Environmental and other phenotype variables
```{r}
# load in second pheno metadata
pheno2 <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/subject_pheno_204.csv")

pheno2 <- pheno2 %>% 
  mutate(across(c(maskid), as.factor))
```
```{r}
# merge pheno type datasets
pheno_complete <- dplyr::left_join(
  pheno_filt_ct_unscaled,
  pheno2,
  by = "maskid")

# clean up duplicate columns
pheno_complete <- pheno_complete %>%
  select(-ends_with(".y")) %>%
  rename_with(~ gsub("\\.x$", "", .x), ends_with(".x"))
```



## Exploratory Data Analysis (EDA)

Methylation Values to Beta function
```{r}
m2beta <- function(m) {
  2^m / (1 + 2^m)}
```

```{r}
# number of patients
cat("Number of patients:",
  length(unique(pheno_filt_ct_unscaled$maskid)),"\n\n")

# number of sample wells
cat("Number of sample Wells:",
  length(unique(pheno_filt_ct_unscaled$Sample_Well)),"\n\n")

# number of sample plates
cat("Number of sample Plates:",
  length(unique(pheno_filt_ct_unscaled$Sample_Plate)),"\n\n")

# number of slides
cat("Number of slides:",
  length(unique(pheno_filt_ct_unscaled$Slide)),"\n\n")

# arrays on slide - 8
unique(pheno_filt_ct_unscaled$Array)
```

```{r}
# number of patients per clinical site ≈73.5% europe
cc_distn <- pheno_filt_ct_unscaled %>% 
  group_by(cc) %>% 
  summarize(counts = n()) %>%
  mutate(percentage = round(counts / sum(counts) * 100, 2))

cc_distn

```


```{r}
# missing value percentages
pheno_filt_ct_unscaled %>%
  ungroup() %>% 
  miss_var_summary()

```

```{r}
# determine number of subjects within age brackets
pheno_filt_ct_unscaled %>%
  mutate(age_group = cut(age_yrs, 
                         breaks = c(0,1,2,3,4,5,6,7), 
                         include.lowest = TRUE)) %>%
  group_by(age_group) %>%
  summarise(n_subjects = n_distinct(maskid)) %>%
  mutate(`%` = round(n_subjects / sum(n_subjects) * 100,2))
  
```

[0,1] = 0 ≤ age ≤ 1
(1,2] = 1 < age ≤ 2


```{r}
pheno_index <- pheno_complete %>%
  mutate(
    age_at_diag = case_endptage / 365.25,
    time_to_index = abs(age_yrs - age_at_diag)) %>%
  group_by(maskid) %>%
  slice_min(time_to_index, with_ties = FALSE) %>%
  ungroup()

pheno_longitudinal <- pheno_filt_ct_unscaled %>%
  group_by(maskid, CaseControl) %>%
  summarise(
    n_visits = n(),
    min_age = min(age_yrs, na.rm = TRUE),
    max_age = max(age_yrs, na.rm = TRUE),
    followup_years = max_age - min_age,
    .groups = "drop")

pheno_subject <- pheno_index %>%
  left_join(pheno_longitudinal, by = "maskid")

pheno_subject$dr34 <- factor(pheno_subject$dr34)
```

### Table 1 Study Characterisitcs
```{r}
tbl1 <- pheno_subject %>%
  mutate(age_at_diag = case_endptage/365.25,
         race = factor(
           na_if(race_ethnicity, 5),
           levels = c(1, 2, 3, 4),
           labels = c("Hispanic (all races)", "NHW", "AA", "Other")),
         fdr_3level = factor(fdr_3level, levels = c("none", "dadsib", "mom"),
                             labels = c("None", "Dad", "Mom")),
         dr34 = case_when(
           is.na(dr34) ~ NA_character_,
           dr34 == 0 ~ "No",
           dr34 == 1 ~ "Yes",
           TRUE ~ NA_character_) |> factor(levels = c("No", "Yes"))) %>% 
  
  tbl_summary(
    by = CaseControl.x,
    include = c(age_at_diag, n_visits, followup_years, dr34, fdr_3level,
                gender, race, cc),
    type = list(
      all_continuous() ~ "continuous2",
      c(dr34, fdr_3level, gender, race, cc) ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    
    digits = all_continuous() ~ 1,
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    
    # change variable names
    label = list(
      gender ~ "Gender",
      race ~ "Race/Ethnicity",
      dr34 ~ "HLA-DR3/DR4",
      fdr_3level ~ "First-Degree Relative",
      age_at_diag  ~ "Age at Diagnosis",
      cc ~ "Clinical Site",
      n_visits ~ "Visits per subject",
      followup_years ~ "Follow-up duration (yrs)"
      )) %>% 

  add_overall() %>%
  bold_labels() %>% 
  italicize_levels() %>% 
  add_n() %>% 
  # modify_spanning_header(
  #   c("stat_1", "stat_2") ~ "****") %>%
  modify_footnote_header(
    footnote = "Cases and controls were matched on age, sex, and clinical site 
    at the index timepoint; therefore, these characteristics are presented for
    the cohort overall rather than contrasted by case-control status",
    columns = all_stat_cols(),
    replace = FALSE) %>% 
  
  as_gt() %>%
  tab_header(
    title = "Table 1. Characteristics of the TEDDY Study Cohort") %>% 
  gt::fmt_markdown(columns = everything()) %>% 
  gt::fmt_number(
    columns = dplyr::starts_with("p.value"),
    decimals = 2) %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    table.width = pct(75),
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))


tbl1
gt::gtsave(tbl1, "/Users/ryan_summers/Desktop/table1.html")
```


```{r}
# measurement stats
pheno_filt_ct_unscaled %>% 
  group_by(maskid) %>%
  summarise(n_measurements = n()) -> mnts_per_id 

mnts_per_id %>%   
  summarise(avg = mean(n_measurements),
            min = min(n_measurements),
            med = median(n_measurements),
            max = max(n_measurements))

hist(mnts_per_id$n_measurements, breaks = 12, freq = FALSE)
```

Number of observations for each participant
```{r}
# determine number of subjects that would give "stable" estimates
subj <- pheno_filt_ct_unscaled %>%
  group_by(maskid) %>%
  summarise(
    n_obs = n(),
    n_age_unique = n_distinct(sample_agedys), 
    .groups = "drop")

# cumulative percentages (e.g. % more than 3 measurements=70.1%)
subj %>%
  reframe(
    threshold = 1:max(n_obs),
    `<=== % gt` = sapply(threshold, function(k) {
      mean(n_obs > k) * 100
      }))

```

```{r}
# check that whole-blood samples have variability
covs <- c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sds <- sapply(pheno_filt_ct_unscaled[, covs], sd, na.rm=TRUE)
which(sds == 0 | is.na(sds))
colSums(!is.finite(as.matrix(pheno_filt_ct_unscaled[, covs])))

```

```{r}
vars <- c("age_yrs_c","new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sapply(pheno_filt_ct_scaled[, vars], function(x) sum(!is.finite(x)))

sapply(pheno_filt_ct_unscaled[, c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")],
       function(x) sd(x, na.rm = TRUE))

```

```{r}
# Confirm sample ID matching across pheno and M-matrix
sid <- as.character(pheno_filt_ct_unscaled$rgName)
mean(sid %in% colnames(matrix.filt)) # should be 1
sum(!(sid %in% colnames(matrix.filt))) # should be 0

```

Sanity check: confirm 1:1 sets
* is there a match for each case indicator? Other words, a case & control for each indicator

```{r}
check_sets <- pheno_filt_ct_unscaled %>%
  dplyr::distinct(case_ind, maskid, outcome, case_endptage) %>%
  group_by(case_ind) %>%
  summarise(
    n_ids = n_distinct(maskid),
    n_cases = sum(outcome == 1),
    n_ctrls = sum(outcome == 0),
    endpt_unique = n_distinct(case_endptage),
    .groups = "drop")

table(check_sets$n_ids) # checks number of IDs per Case Indicator

# identifies case indicators that do not have a match
bad_sets <- check_sets %>% 
  filter(n_ids != 2 | n_cases != 1 | endpt_unique != 1)
bad_sets

```


### Spaghetti plots

```{r}
getSpagPlot <- function(CpG, mvals, pheno, anno) {
  # error handling
  stopifnot(CpG %in% rownames(mvals))
  stopifnot(CpG %in% rownames(anno))
  stopifnot("rgName" %in% names(pheno))
  stopifnot(all(c("sample_agedys","maskid") %in% names(pheno)))

  # match matrix column names by rgName in pheno
  samp <- intersect(colnames(mvals), pheno$rgName)
  if (length(samp) == 0) 
    stop("No overlap between colnames(matrix) and pheno$rgName")

  # build plotting data
  toPlot <- pheno[match(samp, pheno$rgName), , drop = F] # match sample names to pheno
  toPlot$CpG_value <- as.numeric(mvals[CpG, samp]) # match CpG & sample name to matrix

  # annotation for title
  a <- anno[CpG, , drop = F]
  title <- paste0(
    CpG, " (", a$UCSC_RefGene_Name, ") ",
    a$chr, ": ", prettyNum(a$pos, big.mark = ","))

  ggplot(toPlot, 
         aes(x = sample_agedys, y = m2beta(CpG_value), group = maskid)) +
    geom_line(alpha = 0.4) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.10)) +
    theme_bw() +
    ylab("Beta Value") +
    labs(title = title) +
    theme(plot.title = element_text(size = 12))
}



#plots <- lapply(df$CpG, getSpagPlot, toPlot = toPlot, anno = anno)
p <- getSpagPlot("cg10631356", 
                 mvals = matrix.filt, 
                 pheno = pheno_filt_ct_unscaled, 
                 anno = anno)
p

```

### CpG Reports

```{r}
# generate a CpG report
rep <- cpg_report(
  "cg10631356", 
  matrix.filt,
  pheno_filt_ct_unscaled,
  anno,
  DNAm_values = "M",
  outlier_method = "mad",
  z_thresh = 3)

# extract CpG values for each subject along w/ pheno data
rep$data

# annotation
rep$anno

# get outliers
rep$outliers %>% arrange(desc(abs(outlier_score)))
```

```{r}
sample_outlier_counts <- outliers %>%
  dplyr::count(sample)


outliers %>% 
  group_by(CpG) %>% 
  summarize(counts = n()) %>% 
  arrange(desc(counts))
```


```{r}
# median per CpG
med <- rowMedians(matrix.filt, na.rm = TRUE)

# MAD per CpG
mad_vals <- rowMads(matrix.filt, constant = 1.4826, na.rm = TRUE)

# robust z-scores
zMAD <- sweep(matrix.filt, 1, med, "-")
zMAD <- sweep(zMAD, 1, mad_vals, "/") # matrix of MAD scores

# outlier matrix
outlier_mat <- abs(zMAD) >= 3 #??
```


```{r}
idx <- which(outlier_mat, arr.ind = TRUE)

outliers <- data.frame(
  CpG = rownames(matrix.filt)[idx[,1]],
  sample = colnames(matrix.filt)[idx[,2]],
  value = matrix.filt[idx],
  zMAD = zMAD[idx])

qsave(outliers, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/MAD_outliers.qs2",
      nthreads = 6)
```

```{r}
# load outliers identified by MAD dataset
mad_outliers <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/MAD_outliers.qs2",nthreads = 6)

nrow(mad_outliers)/1112858208
```

1.25% (~13.9M) of total observations were flagged as outliers

### M-matrix w/out outliers
```{r}
# M-matrix with MAD outliers removed
matrix.clean <- matrix.filt
matrix.clean[outlier_mat] <- NA # setting only specific samples within each Cpg as NA

qs_save(matrix.clean, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.clean.qs2", nthreads = 6)
```

```{r}
# load clean matrix
matrix.clean <- qs_read("/Users/ryan_summers/GitHub/teddy-dnam-analysis/data/processed/matrix.clean.qs2", nthreads = 6)
```


## Sample Modeling Runs

```{r}
# test a specific CpG cg10631356
cg10631356_mx <- matrix.clean["cg10631356", ]
cg10631356_og <- matrix.filt["cg10631356", ]

cg10631356_df <- data.frame(
  rgName = colnames(matrix.clean),
  M_value = as.numeric(cg10631356_mx))

cg10631356_df2 <- data.frame(
  rgName = colnames(matrix.filt),
  M_value = as.numeric(cg10631356_og))

cg10631356_df <- merge(cg10631356_df, pheno_filt_ct_unscaled, by = "rgName")
cg10631356_df2 <- merge(cg10631356_df2, pheno_filt_ct_unscaled, by = "rgName")

cg10631356.model <- lmer(
  M_value ~ age_yrs + gender + cc + CaseControl + new_Bcell + new_CD4T + 
    new_CD8T + new_Mono + new_NK + (age_yrs | maskid),
  data = cg10631356_df)

summary(cg10631356.model)


cg10631356.model2 <- lmer(
  M_value ~ age_yrs + gender + cc + CaseControl + new_Bcell + new_CD4T + 
    new_CD8T + new_Mono + new_NK + (age_yrs | maskid),
  data = cg10631356_df2)

summary(cg10631356.model2)

nobs(cg10631356.model)
nobs(cg10631356.model2)

```


```{r}
# test a small sample
set.seed(2522)
idx <- sample(nrow(matrix.filt), 500)
matrix.sample <- matrix.filt[idx, , drop = F]

pheno_ordered <- pheno_filt_ct_unscaled[match(colnames(matrix.sample), 
                                     pheno_filt_ct_unscaled$rgName), ]
stopifnot(all(!is.na(pheno_ordered$rgName)))
```

Testing singular CpGs produced from lmeR model w/bobyqa optimizer
```{r}
cpgs_to_test <- lmeR_bobyqa_results[lmeR_bobyqa_results$is_singular==TRUE,]$CpG
matrix.sample <- matrix.filt[cpgs_to_test, , drop = F]

pheno_ordered <- pheno_filt_ct_unscaled[match(colnames(matrix.sample), 
                                     pheno_filt_ct_unscaled$rgName), ]
stopifnot(all(!is.na(pheno_ordered$rgName)))
```


### lmeR Run

```{r}
# fit model
t0 <- Sys.time()

# testing different optimizer settings
ctrl <- lme4::lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5)) #1e5 is default

fit_cpg <- function(probe) {
  mod_function_lmer(
    probe = probe,
    matrix = matrix.sample,
    pheno = pheno_filt_ct_unscaled,
    sample_var = "rgName",    
    id_var = "maskid",           
    age_var = "age_yrs",   
    covs = c("gender", "cc", "CaseControl", "new_Bcell", "new_CD4T", 
             "new_CD8T", "new_Mono", "new_NK"),    
    return_blups = FALSE, # keep FALSE for speed
    control = ctrl,
    random_slope = age_var # 1 is intercept only
  )}

# set number of cores
future::plan(future::multicore, workers = 6)
options(future.globals.maxSize = 20 * 1024^3)  # 20 GiBb

test_probes <- rownames(matrix.sample) # dim = 790944x1407
test_res <- future_lapply(test_probes, fit_cpg)
test_res <- data.table::rbindlist(test_res, fill = TRUE)

qsave(
  test_res, 
  "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_int_only_singular.qs")

t1 <- Sys.time()

message("lmeR runtime: ", 
        round(as.numeric(difftime(t1, t0, units = "mins")), 2), 
        " minutes")
```

```{r}
nrow(test_res %>% filter(!is.na(warn_message)))
nrow(test_res %>% filter(!is.na(error_message)))
```

"Model failed to converge with max|grad| = 0.00399344 (tol = 0.002, component 1)":

That warning means lmer returned estimates, but the optimizer stopped with a gradient that’s a bit larger than the default tolerance. In plain terms:

* the algorithm got close to the optimum, but not close enough to declare full convergence.

So the results aren’t automatically “wrong,” but you should treat that CpG’s fit as potentially unreliable (especially SEs / variance components).


Eigenvalue: 
0.0345 - Substantial random-effects variability
5×10⁻⁹- Essentially zero variability

Interpretation:
There is strong subject-to-subject variability in one direction (intercept), and almost none in the other (slope-related direction).

* Eigenvalues > 0 -> matrix is positive definite (good)
* Very small eigenvalues -> one direction has almost no variance
*	≈ 0 eigenvalue -> model is near-singular

```{r}
mod_function_lmer_traj <- function(
    probe,
    matrix,
    pheno,
    sample_var,
    id_var,
    age_var,
    covs = NULL,
    age_mode = c("linear", "ns"),
    age_df = 3,
    return_predictions = T,
    pred_age_grid = NULL,
    use_lmerTest = F,
    REML = F,
    control = NULL,
    random_slope = F) 
{
  
  age_mode <- match.arg(age_mode)
  
  # ---- checks ----
  if (!is.character(probe) || length(probe) != 1) {
    return(data.frame(
      CpG = NA_character_,
      fit_status = "error",
      error_message = "Probe must be a single string",
      stringsAsFactors = F))
  }
  
  if (!(probe %in% rownames(matrix))) {
    return(data.frame(
      CpG = probe,
      fit_status = "error",
      error_message = "Probe not found in matrix",
      stringsAsFactors = F))
  }
  
  # ---- assemble data ----
  cpg <- as.numeric(matrix[probe, ])
  names(cpg) <- colnames(matrix)
  
  sid_chr <- as.character(pheno[[sample_var]])
  missing_sids <- setdiff(unique(sid_chr), colnames(matrix))
  if (length(missing_sids) > 0) {
    return(data.frame(
      CpG = probe,
      fit_status = "error",
      error_message = paste0("Sample IDs in pheno not found in matrix: ", 
                             length(missing_sids)), stringsAsFactors = F))
  }
  
  pheno$CpG <- cpg[sid_chr]
  
  needed <- c("CpG", id_var, age_var, covs)
  needed <- needed[!is.na(needed)]
  forAnalysis <- pheno[stats::complete.cases(pheno[, needed, drop = F]), ,
                       drop = F]
  
  if (nrow(forAnalysis) < 5) {
    return(data.frame(
      CpG = probe,
      fit_status = "error",
      error_message = "Too few complete observations after filtering",
      stringsAsFactors = F))
  }
  
  forAnalysis[[id_var]] <- as.factor(forAnalysis[[id_var]])
  
  # ---- build age term ----
  age_term <- switch(
    age_mode,
    linear = age_var,
    ns = paste0("splines::ns(", age_var, ", df = ", age_df, ")"))
  
  fixed_terms <- c(age_term, covs)
  fixed_terms <- fixed_terms[nzchar(fixed_terms)]
  fixed_rhs <- paste(fixed_terms, collapse = " + ")
  
  rand_term <- if (isTRUE(random_slope)) {
    paste0("(", age_var, " | ", id_var, ")")
  } else {
    paste0("(1 | ", id_var, ")")
  }
  
  form <- stats::as.formula(
    paste0("CpG ~ ", fixed_rhs, " + ", rand_term))
  
  fit_status <- "ok"
  warn_message <- NA_character_
  error_message <- NA_character_
  
  fit_fun <- function() {
    if (use_lmerTest && requireNamespace("lmerTest", quietly = TRUE)) {
      lmerTest::lmer(form, data = forAnalysis, REML = REML, control = control)
    } else {
      lme4::lmer(form, data = forAnalysis, REML = REML, control = control)
    }
  }
  
  model <- tryCatch(
    withCallingHandlers(
      fit_fun(),
      warning = function(w) {
        if (identical(fit_status, "ok")) {
          fit_status <<- "warning"
          warn_message <<- conditionMessage(w)
        }
        invokeRestart("muffleWarning")
      }),
    error = function(e) {
      fit_status <<- "error"
      error_message <<- conditionMessage(e)
      return(NULL)
    })
  
  if (is.null(model)) {
    return(data.frame(
      CpG = probe,
      fit_status = fit_status,
      warn_message = warn_message,
      error_message = error_message,
      stringsAsFactors = F))
  }
  
  # ---- diagnostics ----
  out <- data.frame(
    CpG = probe,
    fit_status = fit_status,
    warn_message = warn_message,
    error_message = error_message,
    age_mode = age_mode,
    age_df = ifelse(age_mode == "ns", age_df, NA),
    random_slope = random_slope,
    model_call = paste(deparse(stats::formula(model)), collapse = " "),
    nObs = stats::nobs(model),
    AIC = stats::AIC(model),
    BIC = stats::BIC(model),
    logLik = as.numeric(stats::logLik(model)),
    neg2LL = -2 * as.numeric(stats::logLik(model)),
    is_singular = lme4::isSingular(model, tol = 1e-5),
    stringsAsFactors = F)
  
  # ---- optional trajectory predictions ----
  if (isTRUE(return_predictions)) {
    
    if (is.null(pred_age_grid)) {
      pred_age_grid <- seq(
        min(forAnalysis[[age_var]], na.rm = T),
        max(forAnalysis[[age_var]], na.rm = T),
        length.out = 10)
    }
    
    # hold nuisance covariates fixed for population trajectory
    pred_df <- data.frame(age_tmp = pred_age_grid)
    names(pred_df)[1] <- age_var
    
    for (cv in covs) {
      x <- forAnalysis[[cv]]
      if (is.factor(x)) {
        pred_df[[cv]] <- factor(levels(x)[1], levels = levels(x))
      } else {
        pred_df[[cv]] <- stats::median(x, na.rm = TRUE)
      }
    }
    
    # include grouping column for predict; allow population-level prediction
    pred_df[[id_var]] <- forAnalysis[[id_var]][1]
    
    preds <- tryCatch(
      stats::predict(model, newdata = pred_df, re.form = NA, allow.new.levels = TRUE),
      error = function(e) rep(NA_real_, length(pred_age_grid)))
    
    pred_names <- paste0("pred_age_", format(round(pred_age_grid, 3), trim = TRUE))
    pred_row <- as.data.frame(t(preds))
    names(pred_row) <- pred_names
    
    out <- cbind(out, pred_row)
  }
  
  out
}
```


```{r}
compare_age_models_one_cpg <- function(
    probe,
    matrix,
    pheno,
    sample_var,
    id_var,
    age_var,
    covs = NULL,
    control = NULL,
    pred_age_grid = NULL,
    random_slope = FALSE) 
  {
  
  fit_model <- function(age_mode, age_df = NULL) {
    mod_function_lmer_traj(
      probe = probe,
      matrix = matrix,
      pheno = pheno,
      sample_var = sample_var,
      id_var = id_var,
      age_var = age_var,
      covs = covs,
      age_mode = age_mode,
      age_df = ifelse(is.null(age_df), 3, age_df),
      return_predictions = TRUE,
      pred_age_grid = pred_age_grid,
      REML = FALSE,
      control = control,
      random_slope = random_slope)
    }
  
  # fit summary rows
  res_linear <- fit_model("linear")
  res_ns3    <- fit_model("ns", 3)
  res_ns4    <- fit_model("ns", 4)
  
  # refit actual model objects for LRT
  build_data <- function() {
    cpg <- as.numeric(matrix[probe, ])
    names(cpg) <- colnames(matrix)
    ph <- pheno
    ph$CpG <- cpg[as.character(ph[[sample_var]])]
    needed <- c("CpG", id_var, age_var, covs)
    needed <- needed[!is.na(needed)]
    ph <- ph[stats::complete.cases(ph[, needed, drop = FALSE]), , drop = FALSE]
    ph[[id_var]] <- as.factor(ph[[id_var]])
    ph
  }
  
  dat <- build_data()
  
  rand_term <- if (random_slope) {
    paste0("(", age_var, " | ", id_var, ")")
  } else {
    paste0("(1 | ", id_var, ")")
  }
  
  cov_txt <- if (length(covs)) paste("+", paste(covs, collapse = " + ")) else ""
  
  f_linear <- stats::as.formula(
    paste0("CpG ~ ", age_var, " ", cov_txt, " + ", rand_term))
  
  f_ns3 <- stats::as.formula(
    paste0("CpG ~ splines::ns(", age_var, ", df = 3) ", cov_txt, " + ", rand_term))
  
  f_ns4 <- stats::as.formula(
    paste0("CpG ~ splines::ns(", age_var, ", df = 4) ", cov_txt, " + ", rand_term))
  
  m_linear <- tryCatch(lme4::lmer(f_linear, data = dat, REML = FALSE, 
                                  control = control), error = function(e) NULL)
  m_ns3    <- tryCatch(lme4::lmer(f_ns3,    data = dat, REML = FALSE, 
                                  control = control), error = function(e) NULL)
  m_ns4    <- tryCatch(lme4::lmer(f_ns4,    data = dat, REML = FALSE, 
                                  control = control), error = function(e) NULL)
  
  lrt_linear_vs_ns3 <- NA
  lrt_ns3_vs_ns4 <- NA
  
  if (!is.null(m_linear) && !is.null(m_ns3)) {
    lrt_linear_vs_ns3 <- tryCatch(anova(m_linear, m_ns3)$`Pr(>Chisq)`[2],
                                  error = function(e) NA_real_)
  }
  
  if (!is.null(m_ns3) && !is.null(m_ns4)) {
    lrt_ns3_vs_ns4 <- tryCatch(anova(m_ns3, m_ns4)$`Pr(>Chisq)`[2], 
                               error = function(e) NA_real_)
  }
  
  bind_rows <- function(...) {
    data.table::rbindlist(list(...), fill = TRUE)
  }
  
  out <- bind_rows(res_linear, res_ns3, res_ns4)
  out$lrt_linear_vs_ns3 <- lrt_linear_vs_ns3
  out$lrt_ns3_vs_ns4 <- lrt_ns3_vs_ns4
  out
}
```


```{r}
t0 <- Sys.time()

ctrl <- lme4::lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5))

set.seed(2522)
benchmark_probes <- sample(rownames(matrix.clean), 1000)

# grid based on age in years
age_grid <- c(
  0.25, 0.5, 0.75, 1.0,
  1.25, 1.5, 1.75, 2.0,
  2.5, 3.5, 5.0, 6.0)

age_grid <- seq(
  min(pheno_filt_ct$age_yrs, na.rm = TRUE),
  max(pheno_filt_ct$age_yrs, na.rm = TRUE),
  length.out = 12)

pheno_complete <- pheno_complete %>% 
  mutate(across(c(maskid, cc, gender, CaseControl), as.factor))


# set number of cores
future::plan(future::multicore, workers = 6)
options(future.globals.maxSize = 20 * 1024^3)  # 20 GiBb

benchmark_res <- future_lapply(benchmark_probes, function(p) {
  compare_age_models_one_cpg(
    probe = p,
    matrix = matrix.clean,
    pheno = pheno_filt_ct_unscaled,
    sample_var = "rgName",
    id_var = "maskid",
    age_var = "age_yrs",
    covs = c("gender", "cc", "CaseControl", "new_Bcell", "new_CD4T",
             "new_CD8T", "new_Mono", "new_NK"),
    control = ctrl,
    pred_age_grid = age_grid,
    random_slope = T)
  })

benchmark_res <- data.table::rbindlist(benchmark_res, fill = TRUE)
#qsave(benchmark_res, "/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/benchmark_res.qs")

t1 <- Sys.time()
message("lmeR runtime: ", 
        round(as.numeric(difftime(t1, t0, units = "mins")), 2), 
        " minutes")
```

```{r}
benchmark_summary <- benchmark_res %>%
  group_by(age_mode, age_df) %>%
  summarise(
    n = n(),
    mean_AIC = mean(AIC, na.rm = TRUE),
    mean_BIC = mean(BIC, na.rm = TRUE),
    prop_warning = mean(fit_status == "warning", na.rm = TRUE),
    prop_error = mean(fit_status == "error", na.rm = TRUE),
    prop_singular = mean(is_singular, na.rm = TRUE),
    .groups = "drop")

print(benchmark_summary)
```

```{r}
per_cpg_comp <- benchmark_res %>%
  select(CpG, age_mode, age_df, AIC, BIC, lrt_linear_vs_ns3, lrt_ns3_vs_ns4) %>%
  distinct()

# just one row per CpG for the LRT summaries
lrt_summary <- per_cpg_comp %>%
  group_by(CpG) %>%
  summarise(
    lrt_linear_vs_ns3 = max(lrt_linear_vs_ns3, na.rm = TRUE),
    lrt_ns3_vs_ns4 = max(lrt_ns3_vs_ns4, na.rm = TRUE),
    .groups = "drop") %>%
  summarise(
    prop_ns3_better_than_linear_p_lt_0_05 = mean(lrt_linear_vs_ns3 < 0.05, na.rm = TRUE),
    prop_ns4_better_than_ns3_p_lt_0_05 = mean(lrt_ns3_vs_ns4 < 0.05, na.rm = TRUE))

print(lrt_summary)
```

```{r}
# AIC percentage
aic_pct <- per_cpg_comp %>%
  mutate(age_mode_df = ifelse(age_mode == "linear", 
                         "linear", 
                         paste0(age_mode, "_", age_df))) %>% 
  select(CpG, age_mode_df, AIC) %>%
  distinct() %>%
  pivot_wider(names_from = age_mode_df, values_from = AIC) %>%
  summarise(pct = mean(ns_4 < linear, na.rm = TRUE) * 100) %>%
  pull(pct)

# AIC percentage
aic_pct2 <- benchmark_res %>%
    mutate(age_mode_df = ifelse(age_mode == "linear", 
                         "linear", 
                         paste0(age_mode, "_", age_df))) %>% 
  select(CpG, age_mode_df, AIC) %>%
  distinct() %>%
  pivot_wider(names_from = age_mode_df, values_from = AIC) %>%
  summarise(pct = mean(ns_3 < linear, na.rm = TRUE) * 100) %>%
  pull(pct)

# LRT - will always be better - more complexity better model fit

aic_pct
aic_pct2
```

```{r}
# results should be in long format with at least:
# CpG, model, AIC
# where model is something like "linear" and "ns3"

aic_change_summary <- benchmark_res %>%
    mutate(age_mode_df = ifelse(age_mode == "linear", 
                         "linear", 
                         paste0(age_mode, "_", age_df))) %>% 
  select(CpG, age_mode_df, AIC) %>%
  distinct() %>%
  pivot_wider(names_from = age_mode_df, values_from = AIC) %>%
  mutate(
    delta_AIC = linear - ns_3,
    ns3_better = delta_AIC > 0,
    ns3_better_2 = delta_AIC > 2,
    ns3_better_10 = delta_AIC > 10) %>%
  summarise(
    n_cpg = n(),
    pct_ns3_better = mean(ns3_better, na.rm = TRUE) * 100,
    pct_delta_AIC_gt_2 = mean(ns3_better_2, na.rm = TRUE) * 100,
    pct_delta_AIC_gt_10 = mean(ns3_better_10, na.rm = TRUE) * 100,
    mean_delta_AIC = mean(delta_AIC, na.rm = TRUE),
    median_delta_AIC = median(delta_AIC, na.rm = TRUE),
    q25_delta_AIC = quantile(delta_AIC, 0.25, na.rm = TRUE),
    q75_delta_AIC = quantile(delta_AIC, 0.75, na.rm = TRUE),
    q90_delta_AIC = quantile(delta_AIC, 0.90, na.rm = TRUE),
    q95_delta_AIC = quantile(delta_AIC, 0.95, na.rm = TRUE),
    max_delta_AIC = max(delta_AIC, na.rm = TRUE))

print(t(aic_change_summary))
```

Based on median, for more than half of CpGs, linear is slightly better or essentially equivalent.
But, a subset of CpGs are strongly nonlinear:
q90 = 46  
q95 = 94  
max = 911

~38% show real improvement of AIC greater than 2
~22% show strong improvement of AIC greater than 10

So, approximately 

```{r}
aic_change_tbl <- benchmark_res %>%
    mutate(age_mode_df = ifelse(age_mode == "linear", 
                         "linear", 
                         paste0(age_mode, "_", age_df))) %>% 
  select(CpG, age_mode_df, AIC) %>%
  distinct() %>%
  pivot_wider(names_from = age_mode_df, values_from = AIC) %>%
  mutate(
    delta_AIC = linear - ns_3,
    ns3_better = delta_AIC > 0,
    ns3_better_2 = delta_AIC > 2,
    ns3_better_10 = delta_AIC > 10)

head(aic_change_tbl)
```

```{r}
hist(
  aic_change_tbl$delta_AIC,
  breaks = 100,
  main = "Delta AIC: linear - ns3",
  xlab = "Delta AIC")
abline(v = c(0, 2, 10), lty = 2)
```

```{r}
traj_cols <- grep("^pred_age_", names(benchmark_res), value = TRUE)

traj_mat <- as.matrix(benchmark_res[, ..traj_cols])
rownames(traj_mat) <- benchmark_res$CpG

```

```{r}
# split models
dt_linear <- benchmark_res[age_mode == "linear"]
dt_ns3    <- benchmark_res[age_mode == "ns" & age_df == 3]

# ensure correct order
setkey(dt_linear, CpG)
setkey(dt_ns3, CpG)

all(dt_linear$CpG == dt_ns3$CpG)
```

```{r}
# create matrices
traj_linear <- as.matrix(dt_linear[, ..traj_cols])
traj_ns3    <- as.matrix(dt_ns3[, ..traj_cols])
```

```{r}
cor(as.vector(dist(traj_linear)), as.vector(dist(traj_ns3)))
```

Main takeaways:

Group 1 (~50%)
	* mostly linear
	* spline ≈ linear

Group 2 (~35%)
	* moderately nonlinear
	* pline changes curvature but not drastically

Group 3 (~15–20%)
	* strongly nonlinear
	* spline fundamentally changes trajectory
	
Quick and dirty clustering
```{r}
scale_rows <- function(x) {
  t(scale(t(x), center = TRUE, scale = TRUE))
}

traj_linear_scaled <- scale_rows(traj_linear)
traj_ns3_scaled <- scale_rows(traj_ns3)

# handle any constant rows (sd = 0)
traj_linear_scaled[is.na(traj_linear_scaled)] <- 0
traj_ns3_scaled[is.na(traj_ns3_scaled)] <- 0
```

```{r}
km_linear <- kmeans(traj_linear_scaled, centers = 20)
km_ns3 <- kmeans(traj_ns3_scaled, centers = 20)

library(mclust)
adjustedRandIndex(km_linear$cluster, km_ns3$cluster)
```

The Adjusted Rand Index ranges:
	*	1.0 → identical clustering
	*	~0.8–1.0 → very similar
	*	~0.5–0.7 → moderately similar
	*	~0.2–0.4 → weak similarity
	*	~0.0 → essentially unrelated

Check how many subjects only have one observation, which can be ill-posed for
random intercept/slope models

```{r}
pheno_filt_ct_unscaled$maskid <- factor(pheno_filt_ct_unscaled$maskid)
table_sizes <- table(pheno_filt_ct_unscaled$maskid)
summary(table_sizes)
mean(table_sizes == 1)

```

Look at effect sizes

* logFC is the estimated regression coefficient = M-values
* AveExpr is the average level of M-values for CpG site across all samples
```{r}
tt <- topTable(dream.model2, coef="age_yrs_c", number=Inf)
summary(round(m2beta(abs(tt$logFC)), 4))  # for methylation, interpret appropriately
quantile(round(m2beta(abs(tt$logFC)),4), c(.5,.9,.99))

```

```{r}
# genomic inflation factor
chisq <- qchisq(1 - p, df=1)
lambda <- median(chisq, na.rm=TRUE) / qchisq(0.5, df=1)
lambda

```


The below Emprirical Bayes method has been shown to be inefficient:

"We note that an earlier version (before v1.15.5 from September 5, 2019) of dream in the
variancePartition package included an empirical Bayes step. However, after additional
testing, we removed this step so that only gene-level estimates are used in the linear mixed
model. We note that the elevated false positive rate that Yu et al. (2019) observed in simu-
lations using dream are due this this issue and has been resolved." - Hoffman et al 2020 supp
```{r}
# Empirical Bayes - pulls info across all CpGs (stabilizes variances) 
dream.model.EB <- eBayes(dream.model)

# qs methods - soon to be deprecated
qsave(dream.model.EB, "dream.model.EB.qs", preset = "fast", nthreads = 6)


# load dream object
system.time({
  dream.model.EB <- qread("dream.model.EB.qs", nthreads = 8)
  })

```


“For each CpG we attempted a random-slope model. When fits were singular or showed convergence issues, we refit using a reduced random-effects structure (uncorrelated slope or random-intercept only). CpGs were excluded only if the reduced model failed or had insufficient observations.”


```{r}
cg25271479_df <- colData(se.filt) %>% 
  as.data.frame() %>%
  mutate(methylation = assay(se.filt, "Mvalue")["cg25271479", ])

## There is no exact, universally accepted denominator df because FE estimates
## depend on RE str and var components
## Therefore, lmer() omits p-values because df are ambiguous; in practice, lmerTest (Satterthwaite) is the standard way to obtain them, especially for large-scale longitudinal omics analyses.
cg25271479.lmer.model <- lmer(
  methylation ~ sample_agedys + gender + cc + (sample_agedys | maskid),
  data = cg25271479_df)

summary(cg25271479.lmer.model)

## The REML likelihood depends on the fixed‐effects design matrix
## REML integrates out fixed effects
## If you compare models with different fixed effects, the REML likelihoods are not comparable
ranova(cg25271479.lmer.model)

```



How to see what is singular in your model
```{r}
VarCorr(cg25271479.lmer.model)
```


## Model Results Analysis 

### Model Formulation for linear mixed model (LMM)

* 10 fixed effects
* 3 random effects (intercept, slope, cov(int,slope))
* 1 residual variance $\sigma^2$ (within)
* k = 14

Model:
$$
Y_i
= \beta_0

+ \beta_{\text{age}}(\text{age}_i)
+ \beta_{\text{gender}}(\text{gender}_i)
+ \beta_{\text{cc}}(\text{CC}_i)
+ \beta_{\text{CaseControl}}(\text{CaseControl}_i)
+ \beta_{\text{B}}(\text{Bcell}_i) \\
+ \beta_{\text{CD4}}(\text{CD4T}_i)
+ \beta_{\text{CD8}}(\text{CD8T}_i)
+ \beta_{\text{Mono}}(\text{Mono}_i)
+ \beta_{\text{NK}}(\text{NK}_i) \\
+ b_{0}
+ b_{1}(\text{age}_i)
+ \varepsilon_i
$$

$$
\mathbf y_i \sim N(X_i\beta,\ V_i) \\
\mathbf y_i = X_i\beta + Z_i b_i + \epsilon_i
$$

log-likelihood:
$$
\ell =
-\frac{1}{2}
\left[
\log|V|

+ (y-X\beta)^T V^{-1}(y-X\beta)
+ n \log(2\pi)
  \right]
$$

$$
\hat\beta = (X^\top V^{-1} X)^{-1}X^\top V^{-1} y
$$

Subject-level
$$
V_i = Var(Y_i) = Z_iGZ^T_i + R_i
$$

Random effects Var-Cov Matrix
$$
G=
\begin{pmatrix}
\sigma^2_{intercept} & \sigma_{int,slope}\\
\sigma_{int,slope} & \sigma^2_{slope}
\end{pmatrix}
$$

$$
x^T G x = \mathrm{Var}(x^T b) \ge 0
$$
* Strictly positive unless there is perfect linear dependence. I.e symmetric positive-definite (SPD)
* In linear mixed models, covariance matrices are symmetric positive-definite (or semidefinite), and the marginal covariance $V = Z G Z^T + \sigma^2 I$ is SPD when residual variance is positive.
    * $G$ describes variability of random effects
    * $Z D Z^T$ induces correlation between observations
    * $\sigma^2 I$ adds independent noise
    * total covariance must represent a real variance -> SPD
    * if SPD $G = LL^T$ - where $L$ is lower triangular : $\text{det}(G) = (\prod L_{ii})^2 = \prod_i \lambda_i$
      * $\text{det}(G)$ represents total variability of the random effects
      * large determinant -> random effects spread widely
	    * small determinant -> random effects concentrated
	    * determinant = 0 -> ellipsoid collapses -> singular covariance


### Load Results
```{r}
# load in Step1B original model results
step1b_og_results <- qread("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/step1b_og_results.qs", nthreads = 6)
```

```{r}
step1b_scaled_results <- qs_read("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/Step1B_Model_scaled.qs2", nthreads = 6)
```

```{r}
# load lme modeling results from Alpine
lme_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lme_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results from Alpine
lmeR_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results using bobyqa optimizer from Alpine
lmeR_bobyqa_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_bobyqa_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results using bobyqa optimizer from Alpine and w/ BLUPs
lmeR_bobyqa_BLUP_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_bobyqs_wBLUP_results.csv") %>% select(-1)
```

```{r}
# load lmeR modeling results using bobyqa optimizer from Alpine and w/ MAD outliers removed
lmeR_MAD_results <- read_csv("/Users/ryan_summers/GitHub/teddy-dnam-analysis/results/lmeR_MAD_results.csv") %>% select(-1)
```

### Exploration Convergence Issues

```{r}
length(lme_results$var.rand.slope < 1e-8) # should matrix M-matrix
mean(lme_results$var.rand.slope < 1e-8, na.rm=TRUE)
summary(lme_results$var.rand.slope)

```

~5.3% of CpGs have effectively zero random‐slope variance
* there is no between‐subject heterogeneity in age slope

```{r}
# check data matrix (X) for redundancy - DOES NOT EVALUATE Z
mv <- ~ age_yrs_c + gender + cc + new_Bcell + new_CD4T + new_CD8T + new_Mono + new_NK
X <- model.matrix(mv, data = pheno_filt_ct_scaled)

qr(X)$rank
ncol(X)
kappa(X)

```

* Full rank -> no exact collinearity?
* How close the matrix is to being singular
  * Large singular value -> strong direction in predictor space
  * Small singular value -> nearly redundant / weakly identifiable direction
  
It does not say anything about:
* random-effects variance structure
* boundary variance estimates
* random slope identifiability
* covariance singularities

This is why kappa looks good, and we still get convergence warnings. They come from the random-effects (Z), not from X

```{r}
# check that whole-blood samples have variability
covs <- c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sds <- sapply(pheno_filt_ct_unscaled[, covs], sd, na.rm=TRUE)
which(sds == 0 | is.na(sds))
colSums(!is.finite(as.matrix(pheno_filt_ct_unscaled[, covs])))

```

```{r}
vars <- c("sample_agedys","new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")
sapply(pheno_filt_ct_unscaled[, vars], function(x) sum(!is.finite(x)))

sapply(pheno_filt_ct_unscaled[, c("new_Bcell","new_CD4T","new_CD8T","new_Mono","new_NK")],
       function(x) sd(x, na.rm = TRUE))

```



### Identify CpGs that did not converge

```{r}
# identify CpGs not fit in Step1B model
step1b_bad_cpgs <- step1b_og_results %>% filter(!is.na(message))
```

```{r}
# Step1B with scaled variables - CpGs that did not fit
step1b_scaled_bad_cpgs <- step1b_scaled_results %>% filter(!is.na(message))
```

```{r}
# lme model CpGs that did not fit (my script)
lme_bad_cpgs <- lme_results %>% filter(!is.na(error_message))
```

```{r}
# lmeR model CpGs that did not fit (my script)
## the lmeR provides estimates for every CpG but comes w/ convergence warnings
lmeR_bad_cpgs <- lmeR_results %>% filter(!is.na(warn_message))
```

```{r}
# lmeR model, using bobyqa optimizer, CpGs that did not fit
## the lmeR provides estimates for every CpG but comes w/ convergence warnings
lmeR_bobyqa_bad_cpgs <- lmeR_bobyqa_results %>% filter(!is.na(warn_message))
```

```{r}
# how many of the "bad" CpGs overlap between the models
cat("Step1B, lme, lmeR, and lmeR w/bobyqa had", 
    length(intersect(step1b_bad_cpgs$CpG, lme_bad_cpgs$CpG, 
                     lmeR_bad_cpgs, lmeR_bobyqa_bad_cpgs)), 
    "overlapping CpGs that failed\n")

# unique CpGs to all models
unique_step1b <- setdiff( # finds all rows in x that aren't in y : setdiff(x,y)
  step1b_bad_cpgs$CpG,
  lme_bad_cpgs$CpG)

cat("Step1B had", length(unique_step1b), "unique CpGs that failed\n")

unique_lme <- setdiff(
  lme_bad_cpgs$CpG,
  step1b_bad_cpgs$CpG)

cat("LME had", length(unique_lme), "unique CpGs that failed\n")

```

lme runtime: 406.77 minutes ~ 6.8hrs
dream runtime: 219.8 minutes ~ 3.66hrs

Number of singular fits using lmeR
```{r}
cat("lmeR model w/ default optimization had",
    nrow(lmeR_results[lmeR_results$is_singular==TRUE,]),
    "singular fits\n\n")

cat("lmeR model w/ bobyqa optimization had",
    nrow(lmeR_bobyqa_results[lmeR_bobyqa_results$is_singular==TRUE,]),
    "singular fits\n")
```


Compare number of significant slopes across methods before multiple-testing adjustment
```{r}

# Step1B OG
cat("Step1B model had", 
    nrow(subset(step1b_og_results, fixed.sample_agedys.p.value < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(step1b_og_results, fixed.sample_agedys.p.value < 0.05)) / 
      nrow(step1b_og_results),3), "% of total CpG tested\n\n")


# lme - mine
cat("lme model had", 
    nrow(subset(lme_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lme_results, p_slope < 0.05)) / 
      nrow(lme_results),3), "% of total CpG tested\n\n")

# lmeR 
cat("lmeR model had", 
    nrow(subset(lmeR_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lmeR_results, p_slope < 0.05)) / 
      nrow(lmeR_results),3), "% of total CpG tested\n\n")

# lmeR w/ bobyqa optimization
cat("lmeR model w/bobyqa optimization had", 
    nrow(subset(lmeR_bobyqa_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lmeR_bobyqa_results, p_slope < 0.05)) / 
      nrow(lmeR_bobyqa_results),3), "% of total CpG tested\n\n")

# lmeR w/ bobyqa optimization and outliers removed
cat("lmeR model w/bobyqa optimization & MAD outliers removed had", 
    nrow(subset(lmeR_MAD_results, p_slope < 0.05)), 
    "significant slopes representing", 
    100*round(nrow(subset(lmeR_MAD_results, p_slope < 0.05)) / 
      nrow(lmeR_MAD_results),3), "% of total CpG tested\n")

```

* Model w/BOBYQA and MAD outliers removed retained 14450 more CpGs

Identify CpGs that had a significant slope but was singular - var(slope)≈0
```{r}
lmeR_bobyqa_results <- lmeR_bobyqa_results %>% 
  mutate(sig_singular = ifelse(
    is_singular == TRUE & p_slope<0.05, "Trouble", "Good"))

lmeR_results <- lmeR_results %>% 
  mutate(sig_singular = ifelse(
    is_singular == TRUE & p_slope<0.05, "Trouble", "Good"))

lmeR_MAD_results <- lmeR_MAD_results %>% 
  mutate(sig_singular = ifelse(
    is_singular == TRUE & p_slope<0.05, "Trouble", "Good"))


# number of CpGs with a significant slope but was singular
cat("lmeR w/default optimization had",
    nrow(lmeR_results[lmeR_results$sig_singular=="Trouble",]),
    "CpGs with a significant slope but was singular\n", 
    "(subjects are inferred to have same slope)\n\n")

cat("lmeR w/bobyqa optimization had",
    nrow(lmeR_bobyqa_results[lmeR_bobyqa_results$sig_singular=="Trouble",]),
    "CpGs with a significant slope but was singular\n", 
    "(subjects are inferred to have same slope)\n\n")

cat("lmeR w/bobyqa optimization and MAD outliers removed had ",
    nrow(lmeR_bobyqa_results[lmeR_MAD_results$sig_singular=="Trouble",]),
    "CpGs with a significant slope but was singular\n", 
    "(subjects are inferred to have same slope)")
```

* 9474 more CpGs with significant slopes and singular

Using the bobyqa optimization and increasing maxfun let the optimizer converge more fully, and for many CpGs the REML estimate for the random-slope variance is on the boundary (≈0), so more fits are flagged as singular. This suggests the random slope is not needed (no heterogeneity) for those CpGs, and AIC-based screening should be used with caution and looked at explicitly.


### Multiple Testing Adjustment

```{r}
head(lmeR_bobyqa_results)
```

Interpretation for CpG cg26928153:
***Must know the baseline methylation value.***
$$
\frac{d\beta}{dM} = \beta(1-\beta) \text{ln}(2) \\
\frac{d\beta}{dage} = \beta(1-\beta) \text{ln}(2) \times \frac{dM}{dage} \text{(our model estimate)}
$$
Say baseline β is 0.5:
* β=0.5: β(1-β) = 0.25 * ln(2) * β1 = 0.25 x ln(2) x 0.000591 = 0.000102
* β=0.2: β(1-β) = 0.16 * ln(2) * β1 = 0.16 x ln(2) x 0.000591 = 0.000065
* β=0.9: β(1-β) = 0.09 * ln(2) * β1 = 0.09 x ln(2) x 0.000591 = 0.000037

**CpGs near 0.5 -> biggest β change**
**CpGs near 0 or 1 -> tiny β change**

The fixed slope of 0.00059 in M-values corresponds to an approximate β-value increase of 0.00004–0.00010 per year depending on baseline methylation level, indicating a very gradual increase in methylation

In model form:
* For every 1 year increase in age, the log-odds methylation increases by 0.000591
* OR = $2^{0.000591} = 1.00041$: A 1yr increase in age is associated with approximately a 0.041% increase in methylation odds per year (or a 1.00041-fold change in methylation odds)


lmeR w/bobyqa optimization
```{r}
## Should compare and contrast when removing singular slope fits for MultTest adj -if
## we decide to move along that route

# keep CpGs w/ no singular fits
lmeR_bobyqa_results_filt <- lmeR_bobyqa_results %>% 
  filter(is_singular == TRUE)

lmeR_MAD_results_filt <- lmeR_MAD_results %>% 
  filter(is_singular == TRUE)

# Benjamini-Hochberg FDR 
lmeR_bobyqa_results_filt$q_slope <- p.adjust(lmeR_bobyqa_results_filt$p_slope, 
                                            method = "BH")

lmeR_bobyqa_results$q_slope <- p.adjust(lmeR_bobyqa_results$p_slope, 
                                            method = "BH")

lmeR_MAD_results_filt$q_slope <- p.adjust(lmeR_MAD_results_filt$p_slope, 
                                            method = "BH")

lmeR_MAD_results$q_slope <- p.adjust(lmeR_MAD_results$p_slope, 
                                            method = "BH")
```


```{r}
cat(
  "Non-singular fits w/ FDR (BH) adjustment had",
  nrow(lmeR_bobyqa_results_filt[lmeR_bobyqa_results_filt$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n\n")

cat(
  "Singular fits w/ FDR (BH) adjustment had",
  nrow(lmeR_bobyqa_results[lmeR_bobyqa_results$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n\n")

cat(
  "Non-singular fits w/ FDR (BH) adjustment and MAD outliers removed had",
  nrow(lmeR_MAD_results_filt[lmeR_MAD_results_filt$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n\n")

cat(
  "Singular fits w/ FDR (BH) adjustment and MAD outliers removed had",
  nrow(lmeR_MAD_results[lmeR_MAD_results$q_slope < 0.05, ]),
  "CpGs with a significant slope w/ q-values <0.05\n")
```

* With outliers removed, we retained 9139 more CpGs after FDR (BH) adjustment with a significant slope and wasn't singular

### Comparison of p-values with βeta values and AIC

Let's discover any relationships with our p-values/βetas/AIC. Looking at these together may reveal patterns about signal strength, practical relevance, and model structure.

With singular fits using BOBYQA optimization
```{r}
ggplot(lmeR_bobyqa_results, 
       aes(x = fixed_slope, y = -log10(p_slope), color = aic)) +
  geom_point(alpha = 0.25, size = 0.6) +
  scale_color_viridis_c(name = "AIC") +
  labs(x = "Fixed age slope (M-value scale)", y = "-log10(p-value)") +
  theme_classic()
```





